---
title: "Main Analysis: All Studies"
subtitle: "STAT778 Project — gcKrig Robustness Simulation"
author: "Chengyan (Kurt) Ji"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 4.5
)
```

# Purpose

This document is the consolidated analysis layer for the project.
It loads the fit results from all five completed studies and produces
every table and figure cited in the written report:

- **K13A-GRID**: reproduction across the 18-cell parameter grid.
- **MISSPEC**: 2x2 factorial of marginal misspecification.
- **MISSING**: MCAR missingness at five rates.
- **IRREGULAR**: three spatial layouts (grid, random, clustered).
- **ZIP**: zero-inflation with correct vs misspecified fitting.

The per-study walkthrough Rmds (`k13a_grid_analysis.Rmd`,
`misspec_analysis.Rmd`, etc.) remain the source of truth for
detailed inspection and debugging. This file is the report-facing
summary: every numerical claim and every figure in the report is
generated here, so the report can be regenerated by re-knitting
this single document.

# Setup

```{r load-libraries}
suppressPackageStartupMessages({
  library(gcKrig)
})

source("code/R/logging.R")
source("code/R/io.R")
source("code/R/fit_core.R")

# Output directories for tables and figures
dir.create("visualization/report", recursive = TRUE, showWarnings = FALSE)
dir.create("tables/report",        recursive = TRUE, showWarnings = FALSE)
```

```{r helpers}
# Generic tidier: one row per (scenario, replicate, method, parameter).
# Each study supplies its own extractor for scenario-level metadata
# (e.g., missing rate, layout type) via the `extra_cols` function.
tidy_fit <- function(f, extra_cols = function(cfg) data.frame()) {
  r <- f$results
  cfg <- f$scenario_config
  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!isTRUE(r$converged[i])) {
      rows[[length(rows) + 1]] <- cbind(
        scenario_id = f$scenario_id,
        extra_cols(cfg),
        replicate   = r$replicate[i],
        method      = r$method[i],
        converged   = FALSE,
        runtime_sec = r$runtime_sec[i],
        parameter   = NA_character_,
        estimate    = NA_real_,
        stringsAsFactors = FALSE
      )
      next
    }
    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next
    for (pname in names(est)) {
      rows[[length(rows) + 1]] <- cbind(
        scenario_id = f$scenario_id,
        extra_cols(cfg),
        replicate   = r$replicate[i],
        method      = r$method[i],
        converged   = TRUE,
        runtime_sec = r$runtime_sec[i],
        parameter   = pname,
        estimate    = as.numeric(est[[pname]]),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, rows)
}

# Round a numeric column to fixed digits, returning a character vector
# suitable for inclusion in printed tables. Avoids R's default
# scientific notation surprises in MAE/bias columns.
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)

# Save a data frame as a CSV in tables/report and also return it
# invisibly so chunk output remains usable.
save_table <- function(df, name) {
  path <- file.path("tables/report", paste0(name, ".csv"))
  write.csv(df, path, row.names = FALSE)
  invisible(df)
}
```

# Study 1: K13A-GRID Reproduction

The 18-cell factorial: $n \in \{25, 49\}$, range $\in \{0.1, 0.3, 0.5\}$,
$\lambda \in \{1, 5, 10\}$. 200 replications per cell, both GHK and GQT.
This study anchors the report by establishing how the two methods
behave under well-specified conditions.

## Load and tidy

```{r k13a-load}
DATASET_K13A <- "K13A-GRID"
ids_k13a <- list_scenarios(DATASET_K13A, dir = "data/generated")
fits_k13a <- lapply(ids_k13a, function(id) load_fit_result(DATASET_K13A, id))
names(fits_k13a) <- ids_k13a

# Sanity check: 18 cells.
stopifnot(length(fits_k13a) == 18)

extra_k13a <- function(cfg) data.frame(
  n_total = cfg$layout$n_side^2,
  range   = cfg$corr$range,
  lambda  = cfg$marginal$lambda,
  stringsAsFactors = FALSE
)

long_k13a <- do.call(rbind, lapply(fits_k13a, tidy_fit, extra_cols = extra_k13a))

# Attach truth (Poisson intercept is on log scale).
long_k13a$truth <- NA_real_
is_int <- long_k13a$parameter %in% "Intercept"
is_rng <- long_k13a$parameter %in% "range"
long_k13a$truth[is_int] <- log(long_k13a$lambda[is_int])
long_k13a$truth[is_rng] <- long_k13a$range[is_rng]
long_k13a$abs_err <- abs(long_k13a$estimate - long_k13a$truth)
long_k13a$err     <- long_k13a$estimate - long_k13a$truth
```

## Convergence

```{r k13a-convergence}
conv_k13a <- aggregate(
  converged ~ method,
  data = unique(long_k13a[, c("scenario_id", "replicate", "method", "converged")]),
  FUN  = mean
)
conv_k13a$conv_pct <- 100 * conv_k13a$converged
print(conv_k13a, row.names = FALSE)

save_table(conv_k13a, "k13a_convergence")
```

## Runtime

```{r k13a-runtime}
rt_k13a <- unique(long_k13a[long_k13a$converged,
                            c("scenario_id", "replicate", "method",
                              "n_total", "runtime_sec")])
runtime_k13a <- aggregate(runtime_sec ~ n_total + method, data = rt_k13a, FUN = mean)
runtime_wide <- reshape(runtime_k13a, idvar = "n_total", timevar = "method",
                        direction = "wide")
runtime_wide$ratio_GHK_to_GQT <- runtime_wide$runtime_sec.GHK /
                                 runtime_wide$runtime_sec.GQT
print(runtime_wide, row.names = FALSE, digits = 3)

save_table(runtime_wide, "k13a_runtime")

cat(sprintf(
  "\nOverall GHK-to-GQT speedup: %.2fx (n=25), %.2fx (n=49)\n",
  runtime_wide$ratio_GHK_to_GQT[runtime_wide$n_total == 25],
  runtime_wide$ratio_GHK_to_GQT[runtime_wide$n_total == 49]
))
```

## MAE per cell

```{r k13a-mae}
mae_k13a <- aggregate(
  abs_err ~ n_total + range + lambda + method + parameter,
  data = long_k13a[long_k13a$converged, ],
  FUN  = mean
)
names(mae_k13a)[6] <- "MAE"

# Wide form for direct comparison.
mae_k13a_wide <- reshape(
  mae_k13a,
  idvar = c("n_total", "range", "lambda", "parameter"),
  timevar = "method", direction = "wide"
)
mae_k13a_wide$pct_diff <- 100 * (mae_k13a_wide$MAE.GQT - mae_k13a_wide$MAE.GHK) /
                                 mae_k13a_wide$MAE.GHK

# Sort and display
mae_k13a_wide <- mae_k13a_wide[order(mae_k13a_wide$parameter,
                                     mae_k13a_wide$n_total,
                                     mae_k13a_wide$range,
                                     mae_k13a_wide$lambda), ]
print(mae_k13a_wide, row.names = FALSE, digits = 3)

save_table(mae_k13a_wide, "k13a_mae_wide")

cat(sprintf("\nMax absolute %% MAE difference between methods: %.1f%%\n",
            max(abs(mae_k13a_wide$pct_diff), na.rm = TRUE)))
```

## Bias summary

Bias (mean signed error) per cell. Useful for confirming that the
estimators are approximately unbiased and that any MAE growth with
range or low-lambda is variance-driven rather than bias-driven.

```{r k13a-bias}
bias_k13a <- aggregate(
  err ~ n_total + range + lambda + method + parameter,
  data = long_k13a[long_k13a$converged, ],
  FUN  = mean
)
names(bias_k13a)[6] <- "bias"

# Compact summary: mean and max |bias| per (method, parameter).
bias_compact <- aggregate(
  bias ~ method + parameter, data = bias_k13a,
  FUN = function(x) c(mean = mean(x), max_abs = max(abs(x)))
)
bias_compact <- do.call(data.frame, bias_compact)
print(bias_compact, row.names = FALSE, digits = 3)

save_table(bias_k13a, "k13a_bias")
```

## Numerical summary of MAE vs sample size

```{r k13a-mae-vs-n-table}
# this table is the numerical version of the MAE-vs-n plots.
# each row fixes the parameter, true range, lambda, and method,
# then compares MAE at n = 25 and n = 49.

mae_vs_n <- mae_k13a[, c("parameter", "range", "lambda", "method",
                         "n_total", "MAE")]

mae_vs_n_wide <- reshape(
  mae_vs_n,
  idvar = c("parameter", "range", "lambda", "method"),
  timevar = "n_total",
  direction = "wide"
)

# make column names clearer for reporting
names(mae_vs_n_wide)[names(mae_vs_n_wide) == "range"] <- "true_range"
names(mae_vs_n_wide)[names(mae_vs_n_wide) == "MAE.25"] <- "MAE_n25"
names(mae_vs_n_wide)[names(mae_vs_n_wide) == "MAE.49"] <- "MAE_n49"

# change from n = 25 to n = 49
mae_vs_n_wide$change_49_minus_25 <- mae_vs_n_wide$MAE_n49 -
                                    mae_vs_n_wide$MAE_n25

# positive values mean MAE decreased when n increased
mae_vs_n_wide$pct_reduction <- 100 *
  (mae_vs_n_wide$MAE_n25 - mae_vs_n_wide$MAE_n49) /
  mae_vs_n_wide$MAE_n25

mae_vs_n_wide <- mae_vs_n_wide[
  order(mae_vs_n_wide$parameter,
        mae_vs_n_wide$true_range,
        mae_vs_n_wide$lambda,
        mae_vs_n_wide$method),
]

print(mae_vs_n_wide, row.names = FALSE, digits = 3)

save_table(mae_vs_n_wide, "k13a_mae_vs_n_wide")
```


This gives the exact numerical representation of each line in the plot below.

A compact summary averaging over \(\lambda\):

## Average sample-size effect by true range

```{r k13a-mae-vs-n-by-range}
# average the n = 25 to n = 49 improvement over lambda values.
# this summarizes the sample-size effect within each true range panel.

mae_n_effect_by_range <- aggregate(
  cbind(MAE_n25, MAE_n49, change_49_minus_25, pct_reduction) ~
    parameter + true_range + method,
  data = mae_vs_n_wide,
  FUN = mean
)

mae_n_effect_by_range <- mae_n_effect_by_range[
  order(mae_n_effect_by_range$parameter,
        mae_n_effect_by_range$true_range,
        mae_n_effect_by_range$method),
]

print(mae_n_effect_by_range, row.names = FALSE, digits = 3)

save_table(mae_n_effect_by_range, "k13a_mae_n_effect_by_range")
```


## Headline figure: MAE vs sample size, faceted by range

```{r k13a-fig-mae}
plot_mae_panel <- function(mae_tab, param_name) {
  d <- mae_tab[mae_tab$parameter == param_name, ]
  ranges <- sort(unique(d$range))
  par(mfrow = c(1, length(ranges)),
      oma = c(0, 0, 3, 0), mar = c(4, 4, 3, 1))
  for (r in ranges) {
    dr <- d[d$range == r, ]
    y_max <- max(dr$MAE, na.rm = TRUE) * 1.1
    plot(NA, xlim = range(dr$n_total), ylim = c(0, y_max),
         xlab = "n", ylab = "MAE",
         main = sprintf("range = %g", r))
    lams <- sort(unique(dr$lambda))
    cols <- c("firebrick", "steelblue", "darkgreen")
    ltys <- c(1, 2)
    for (li in seq_along(lams)) {
      for (mi in seq_along(c("GHK", "GQT"))) {
        m <- c("GHK", "GQT")[mi]
        sub <- dr[dr$lambda == lams[li] & dr$method == m, ]
        sub <- sub[order(sub$n_total), ]
        lines(sub$n_total, sub$MAE,
              col = cols[li], lty = ltys[mi], lwd = 2)
        points(sub$n_total, sub$MAE, col = cols[li], pch = 19)
      }
    }
    if (r == ranges[1]) {
      legend("bottomright",
             legend = c(expression(lambda == 1),
                        expression(lambda == 5),
                        expression(lambda == 10),
                        "GHK", "GQT"),
             col = c("firebrick", "steelblue", "darkgreen", "black", "black"),
             lty = c(1, 1, 1, 1, 2),
             lwd = 2,
             pch = c(19, 19, 19, NA, NA),
             bty = "n",
             cex = 1)
    }
  }

  mtext(sprintf("%s: MAE vs n (solid=GHK, dashed=GQT, colors=lambda)",
                param_name),
        outer = TRUE, cex = 1.1)
  par(mfrow = c(1, 1))
}

plot_mae_panel(mae_k13a, "Intercept")

pdf("visualization/report/k13a_mae_intercept.pdf", width = 6, height = 5)
plot_mae_panel(mae_k13a, "Intercept")
dev.off()

plot_mae_panel(mae_k13a, "range")

pdf("visualization/report/k13a_mae_range.pdf", width = 6, height = 5)
plot_mae_panel(mae_k13a, "range")
dev.off()
```

# Study 2: MISSPEC (Marginal Misspecification)

A 2x2 factorial: generating family (Poisson or NegBin) crossed with
fitting family (Poisson or NegBin), $n = 49$, $r = 0.3$, 30 replications
per cell. Tests whether ignoring overdispersion or imposing it
unnecessarily distorts estimation.

## Load fit results

```{r misspec-load}
fit_dir_misspec <- file.path("data", "fits", "MISSPEC")
fit_files_misspec <- list.files(fit_dir_misspec, pattern = "\\.rds$",
                                full.names = TRUE)
stopifnot(length(fit_files_misspec) == 4)
fits_misspec <- lapply(fit_files_misspec, readRDS)
names(fits_misspec) <- basename(fit_files_misspec)
```

## Build long-format table with truth attached

```{r misspec-tidy}
truth_for_misspec <- function(pname, gen_cfg, fit_spec) {
  p <- tolower(gsub("[[:space:]]+", "", pname))
  gen_marg <- gen_cfg$marginal
  if (grepl("range", p))  return(gen_cfg$corr$range)
  if (grepl("kappa", p))  return(gen_cfg$corr$kappa)
  if (grepl("nugget", p)) return(gen_cfg$corr$nugget)
  if (grepl("^od$|overdisp|dispersion", p)) {
    if (gen_marg$family == "negbin" && fit_spec$family == "negbin") {
      return(gen_marg$od)
    }
    return(NA_real_)
  }
  # Intercept-scale mean
  if (gen_marg$family == "poisson") return(log(gen_marg$lambda))
  if (gen_marg$family == "negbin")  return(gen_marg$beta[1])
  NA_real_
}

flatten_misspec <- function(fr) {
  r <- fr$results
  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!isTRUE(r$converged[i])) next
    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next
    for (pname in names(est)) {
      truth_val <- truth_for_misspec(pname, fr$scenario_config,
                                     fr$fit_marginal_spec)
      rows[[length(rows) + 1]] <- data.frame(
        gen_family   = fr$scenario_config$marginal$family,
        fit_family   = fr$fit_marginal_spec$family,
        misspecified = fr$misspecified,
        cell_label   = paste(fr$scenario_config$marginal$family, "->",
                             fr$fit_marginal_spec$family),
        replicate    = r$replicate[i],
        method       = r$method[i],
        runtime_sec  = r$runtime_sec[i],
        loglik       = r$loglik[i],
        parameter    = pname,
        estimate     = as.numeric(est[[pname]]),
        truth        = truth_val,
        abs_err      = if (is.na(truth_val)) NA_real_
                       else abs(as.numeric(est[[pname]]) - truth_val),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, rows)
}

long_misspec <- do.call(rbind, lapply(fits_misspec, flatten_misspec))
```

## Convergence and runtime

```{r misspec-conv-rt}
# Reconstruct one-row-per-fit for convergence purposes.
fit_rows_misspec <- do.call(rbind, lapply(fits_misspec, function(fr) {
  r <- fr$results
  data.frame(
    cell_label = paste(fr$scenario_config$marginal$family, "->",
                       fr$fit_marginal_spec$family),
    method     = r$method,
    converged  = r$converged,
    runtime_sec = r$runtime_sec
  )
}))

conv_misspec <- aggregate(converged ~ cell_label + method,
                          data = fit_rows_misspec, FUN = mean)
print(conv_misspec, row.names = FALSE)

rt_misspec <- aggregate(runtime_sec ~ cell_label + method,
                        data = fit_rows_misspec[fit_rows_misspec$converged, ],
                        FUN = mean)
print(rt_misspec, row.names = FALSE, digits = 3)
```

## Headline MAE table

```{r misspec-mae}
mae_misspec <- aggregate(
  abs_err ~ cell_label + method + parameter,
  data = long_misspec[!is.na(long_misspec$abs_err), ],
  FUN  = mean
)
names(mae_misspec)[4] <- "MAE"
mae_misspec <- mae_misspec[order(mae_misspec$parameter,
                                 mae_misspec$cell_label,
                                 mae_misspec$method), ]
print(mae_misspec, row.names = FALSE, digits = 3)

save_table(mae_misspec, "misspec_mae")

# Estimate summary so we can report range bias under NegBin -> Poisson.
est_summary_misspec <- aggregate(
  estimate ~ cell_label + method + parameter,
  data = long_misspec, FUN = mean
)
names(est_summary_misspec)[4] <- "mean_est"
print(est_summary_misspec[est_summary_misspec$cell_label == "negbin -> poisson", ],
      row.names = FALSE, digits = 3)
```

## GQT log-likelihood pathology under NegBin -> Poisson

The GQT surrogate diverges numerically in this misspecification cell.
We document it explicitly because AIC/BIC computed downstream from
GQT cannot be trusted here even when the point estimates look fine.

```{r misspec-gqt-loglik}
ll_summary <- aggregate(
  loglik ~ cell_label + method,
  data = long_misspec[long_misspec$parameter == "Intercept", ],
  FUN  = mean
)
print(ll_summary, row.names = FALSE)
```

## Boxplot of estimates (the report-ready figure)

```{r misspec-boxplot}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  p_box <- ggplot(long_misspec,
                  aes(x = method, y = estimate, fill = method)) +
    geom_boxplot(outlier.alpha = 0.4) +
    facet_grid(parameter ~ cell_label, scales = "free_y") +
    geom_hline(aes(yintercept = truth), color = "red",
               linetype = 2, na.rm = TRUE) +
    labs(x = NULL, y = "Estimate",
         title = "Estimate distributions across MISSPEC cells") +
    theme_minimal()
  print(p_box)
  ggsave("visualization/report/misspec_estimates.pdf",
         p_box, width = 10, height = 6)
}
```

# Study 3: MISSING (MCAR Robustness)

Five MCAR rates $\{0, 0.1, 0.2, 0.3, 0.5\}$ on the Poisson(5) baseline,
$n = 49$, $r = 0.3$, 200 replications per rate.

## Load and tidy

```{r missing-load}
DATASET_MISS <- "MISSING"
ids_miss <- list_scenarios(DATASET_MISS, dir = "data/generated")
fits_miss <- lapply(ids_miss, function(id) load_fit_result(DATASET_MISS, id))
names(fits_miss) <- ids_miss

extra_miss <- function(cfg) data.frame(
  missing_rate = cfg$missing_rate,
  stringsAsFactors = FALSE
)

long_miss <- do.call(rbind, lapply(fits_miss, tidy_fit, extra_cols = extra_miss))
long_miss$truth <- NA_real_
long_miss$truth[long_miss$parameter %in% "Intercept"] <- log(5)
long_miss$truth[long_miss$parameter %in% "range"]     <- 0.3
long_miss$abs_err <- abs(long_miss$estimate - long_miss$truth)
long_miss$err     <- long_miss$estimate - long_miss$truth
```

## MAE and bias by rate

```{r missing-mae}
mae_miss <- aggregate(
  abs_err ~ missing_rate + method + parameter,
  data = long_miss[long_miss$converged, ],
  FUN  = mean
)
names(mae_miss)[4] <- "MAE"

bias_miss <- aggregate(
  err ~ missing_rate + method + parameter,
  data = long_miss[long_miss$converged, ],
  FUN  = mean
)
names(bias_miss)[4] <- "bias"

mae_miss_wide <- reshape(
  mae_miss, idvar = c("missing_rate", "parameter"),
  timevar = "method", direction = "wide"
)
print(mae_miss_wide, row.names = FALSE, digits = 3)

save_table(mae_miss_wide, "missing_mae")

# Display bias (claim: stays near zero across all rates under MCAR).
bias_miss_wide <- reshape(
  bias_miss, idvar = c("missing_rate", "parameter"),
  timevar = "method", direction = "wide"
)
cat("\nBias by missingness rate:\n")
print(bias_miss_wide, row.names = FALSE, digits = 3)

save_table(bias_miss_wide, "missing_bias")

# Headline: MAE inflation from 0% to 50%
for (param in c("Intercept", "range")) {
  for (m in c("GHK", "GQT")) {
    base <- mae_miss$MAE[mae_miss$missing_rate == 0   &
                         mae_miss$parameter == param  &
                         mae_miss$method == m]
    high <- mae_miss$MAE[mae_miss$missing_rate == 0.5 &
                         mae_miss$parameter == param  &
                         mae_miss$method == m]
    cat(sprintf("%s / %s: MAE 0%% -> 50%% : %.4f -> %.4f (%.2fx)\n",
                param, m, base, high, high / base))
  }
}
```

## Runtime by missingness rate

Smaller effective sample sizes mean fewer likelihood evaluations, so
runtime should decrease with the missingness rate. The report cites
the GHK 0% vs 50% comparison directly.

```{r missing-runtime}
rt_miss <- unique(long_miss[long_miss$converged & long_miss$parameter == "Intercept",
                            c("scenario_id", "replicate", "method",
                              "missing_rate", "runtime_sec")])
runtime_miss <- aggregate(runtime_sec ~ missing_rate + method, data = rt_miss, FUN = mean)
runtime_miss_wide <- reshape(runtime_miss, idvar = "missing_rate",
                             timevar = "method", direction = "wide")
print(runtime_miss_wide, row.names = FALSE, digits = 3)

save_table(runtime_miss_wide, "missing_runtime")
```

## Figure: MAE vs missingness rate

```{r missing-fig}
plot_missing <- function() {
  par(mfrow = c(1, 2))
  for (param in c("Intercept", "range")) {
    d <- mae_miss[mae_miss$parameter == param, ]
    y_max <- max(d$MAE) * 1.1
    plot(NA, xlim = c(0, 0.5), ylim = c(0, y_max),
         xlab = "MCAR missingness rate", ylab = "MAE",
         main = sprintf("MAE vs missingness -- %s", param))
    for (m in c("GHK", "GQT")) {
      sub <- d[d$method == m, ]
      sub <- sub[order(sub$missing_rate), ]
      col <- ifelse(m == "GHK", "firebrick", "steelblue")
      lines(sub$missing_rate, sub$MAE, col = col, lwd = 2)
      points(sub$missing_rate, sub$MAE, col = col, pch = 19, cex = 1.2)
    }
    legend("topleft", legend = c("GHK", "GQT"),
           col = c("firebrick", "steelblue"), lwd = 2, pch = 19, bty = "n")
  }
  par(mfrow = c(1, 1))
}
plot_missing()

pdf("visualization/report/missing_mae_vs_rate.pdf", width = 10, height = 4.5)
plot_missing()
dev.off()
```

# Study 4: IRREGULAR (Spatial Layout)

Three layouts at $n = 49$: regular $7 \times 7$ grid, uniform random, and
four Gaussian clusters. Same generating parameters as MISSING:
Poisson(5), $r = 0.3$, 200 replications per layout.

## Load and tidy

```{r irregular-load}
DATASET_IRR <- "IRREGULAR"
ids_irr <- list_scenarios(DATASET_IRR, dir = "data/generated")
fits_irr <- lapply(ids_irr, function(id) load_fit_result(DATASET_IRR, id))
names(fits_irr) <- ids_irr

extra_irr <- function(cfg) data.frame(
  layout_type = cfg$layout$layout_type,
  stringsAsFactors = FALSE
)

long_irr <- do.call(rbind, lapply(fits_irr, tidy_fit, extra_cols = extra_irr))
long_irr$truth <- NA_real_
long_irr$truth[long_irr$parameter %in% "Intercept"] <- log(5)
long_irr$truth[long_irr$parameter %in% "range"]     <- 0.3
long_irr$abs_err <- abs(long_irr$estimate - long_irr$truth)
long_irr$err     <- long_irr$estimate - long_irr$truth
```

## MAE and bias by layout

```{r irregular-mae}
mae_irr <- aggregate(
  abs_err ~ scenario_id + method + parameter,
  data = long_irr[long_irr$converged, ],
  FUN  = mean
)
names(mae_irr)[4] <- "MAE"
mae_irr <- mae_irr[order(mae_irr$parameter, mae_irr$scenario_id,
                         mae_irr$method), ]
print(mae_irr, row.names = FALSE, digits = 3)

save_table(mae_irr, "irregular_mae")

bias_irr <- aggregate(
  err ~ scenario_id + method + parameter,
  data = long_irr[long_irr$converged, ],
  FUN  = mean
)
names(bias_irr)[4] <- "bias"
print(bias_irr, row.names = FALSE, digits = 3)
```

## Figure: layouts side-by-side, then MAE bars

```{r irregular-layouts-fig}
plot_layouts <- function() {
  par(mfrow = c(1, 3))
  for (id in c("grid", "random", "clustered")) {
    scen <- load_scenario(DATASET_IRR, id, dir = "data/generated")
    plot(scen$locations[, "x"], scen$locations[, "y"],
         pch = 19, cex = 1.0, col = "steelblue",
         xlab = "x", ylab = "y", xlim = c(0, 1), ylim = c(0, 1),
         main = sprintf("%s (n = %d)", id, nrow(scen$locations)))
  }
  par(mfrow = c(1, 1))
}
plot_layouts()

pdf("visualization/report/irregular_layouts.pdf", width = 10, height = 3.5)
plot_layouts()
dev.off()
```

```{r irregular-mae-fig}
plot_irr_mae <- function() {
  par(mfrow = c(1, 2))
  for (param in c("Intercept", "range")) {
    d <- mae_irr[mae_irr$parameter == param, ]
    layouts_order <- c("grid", "random", "clustered")
    d$scenario_id <- factor(d$scenario_id, levels = layouts_order)
    wide <- reshape(d[, c("scenario_id", "method", "MAE")],
                    idvar = "scenario_id", timevar = "method",
                    direction = "wide")
    mat <- rbind(wide$MAE.GHK, wide$MAE.GQT)
    colnames(mat) <- wide$scenario_id
    rownames(mat) <- c("GHK", "GQT")
    barplot(mat, beside = TRUE,
            col = c("firebrick", "steelblue"),
            ylab = "MAE",
            main = sprintf("MAE by layout -- %s", param),
            legend.text = rownames(mat),
            args.legend = list(x = "topright", bty = "n"))
  }
  par(mfrow = c(1, 1))
}
plot_irr_mae()

pdf("visualization/report/irregular_mae.pdf", width = 10, height = 4.5)
plot_irr_mae()
dev.off()
```

# Study 5: ZIP (Zero-Inflation)

Three zero-inflation levels $p_0 \in \{0, 0.2, 0.4\}$, each fitted both
correctly (ZIP) and incorrectly (Poisson). $n = 49$, $r = 0.3$.

## Load both well-specified and misspecified fits

```{r zip-load}
DATASET_ZIP <- "ZIP"
inv_zip <- list_fit_results(DATASET_ZIP, dir = "data/fits")

fits_zip <- lapply(seq_len(nrow(inv_zip)), function(i) {
  load_fit_result(
    DATASET_ZIP,
    inv_zip$scenario_id[i],
    misspec_family = if (is.na(inv_zip$misspec_family[i])) NULL
                     else inv_zip$misspec_family[i]
  )
})
names(fits_zip) <- paste(
  inv_zip$scenario_id,
  ifelse(is.na(inv_zip$misspec_family), "wellspec", inv_zip$misspec_family),
  sep = "__"
)

# Custom tidier because ZIP needs both p0 and the actual fit family.
tidy_zip <- function(f) {
  r <- f$results
  cfg <- f$scenario_config
  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!isTRUE(r$converged[i])) next
    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next
    for (pname in names(est)) {
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id = f$scenario_id,
        p0          = cfg$marginal$od,  # ZIP probability stored under `od`
        fit_family  = f$fit_marginal_spec$family,
        replicate   = r$replicate[i],
        method      = r$method[i],
        parameter   = pname,
        estimate    = as.numeric(est[[pname]]),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, rows)
}

long_zip <- do.call(rbind, lapply(fits_zip, tidy_zip))
long_zip$truth <- NA_real_
long_zip$truth[long_zip$parameter %in% "Intercept"] <- log(5)
long_zip$truth[long_zip$parameter %in% "range"]     <- 0.3
long_zip$abs_err <- abs(long_zip$estimate - long_zip$truth)
```

## Convergence

GQT-ZIP shows mild convergence drops at $p_0 \in \{0.2, 0.4\}$. This is
the first sign of the GQT instability under strong zero-inflation,
which the MAE comparison and well-specified-GQT range error make
explicit.

```{r zip-convergence}
# Reconstruct one-row-per-fit for convergence purposes (the long
# table dropped failed rows during tidying).
conv_zip_rows <- do.call(rbind, lapply(fits_zip, function(f) {
  cfg <- f$scenario_config
  data.frame(
    p0          = cfg$marginal$od,
    fit_family  = f$fit_marginal_spec$family,
    method      = f$results$method,
    converged   = f$results$converged,
    stringsAsFactors = FALSE
  )
}))
conv_zip <- aggregate(converged ~ p0 + fit_family + method,
                      data = conv_zip_rows, FUN = mean)
conv_zip <- conv_zip[order(conv_zip$fit_family, conv_zip$p0, conv_zip$method), ]
print(conv_zip, row.names = FALSE, digits = 4)

save_table(conv_zip, "zip_convergence")
```

## MAE: well-specified vs misspecified, side by side

```{r zip-mae}
mae_zip <- aggregate(
  abs_err ~ p0 + fit_family + method + parameter,
  data = long_zip[!is.na(long_zip$abs_err), ],
  FUN  = mean
)
names(mae_zip)[5] <- "MAE"

mae_zip_wide <- reshape(
  mae_zip, idvar = c("p0", "method", "parameter"),
  timevar = "fit_family", direction = "wide"
)
mae_zip_wide$pct <- 100 * (mae_zip_wide$MAE.poisson - mae_zip_wide$MAE.zip) /
                          mae_zip_wide$MAE.zip
mae_zip_wide <- mae_zip_wide[order(mae_zip_wide$parameter,
                                   mae_zip_wide$p0,
                                   mae_zip_wide$method), ]
print(mae_zip_wide, row.names = FALSE, digits = 3)

save_table(mae_zip_wide, "zip_mae_wide")
```

## Figure: MAE vs zero-inflation level

```{r zip-fig}
plot_zip <- function() {
  par(mfrow = c(1, 2))
  for (param in c("Intercept", "range")) {
    d <- mae_zip[mae_zip$parameter == param, ]
    y_max <- max(d$MAE) * 1.1
    plot(NA, xlim = c(0, 0.45), ylim = c(0, y_max),
         xlab = "zero-inflation probability p0",
         ylab = "MAE",
         main = sprintf("MAE vs p0 -- %s", param))
    for (ff in c("zip", "poisson")) {
      for (m in c("GHK", "GQT")) {
        sub <- d[d$fit_family == ff & d$method == m, ]
        sub <- sub[order(sub$p0), ]
        col <- if (ff == "zip") "darkgreen" else "firebrick"
        lty <- if (m == "GHK") 1 else 2
        lines(sub$p0, sub$MAE, col = col, lty = lty, lwd = 2)
        points(sub$p0, sub$MAE, col = col, pch = 19, cex = 1.2)
      }
    }
    legend("topleft",
           legend = c("ZIP fit (well-spec)", "Poisson fit (misspec)",
                      "GHK (solid)", "GQT (dashed)"),
           col = c("darkgreen", "firebrick", "black", "black"),
           lty = c(1, 1, 1, 2), lwd = 2, bty = "n", cex = 0.8)
  }
  par(mfrow = c(1, 1))
}
plot_zip()

pdf("visualization/report/zip_mae_vs_p0.pdf", width = 10, height = 4.5)
plot_zip()
dev.off()
```

## GQT instability at p0 = 0.4

The most striking single finding: even when ZIP is well-specified, GQT
inflates range MAE at $p_0 = 0.4$ to roughly twice GHK's value.

```{r zip-gqt-instability}
zip_well_p04 <- mae_zip[mae_zip$fit_family == "zip" &
                        mae_zip$p0 == 0.4 &
                        mae_zip$parameter == "range", ]
print(zip_well_p04, row.names = FALSE, digits = 3)
```

# Headline Numbers for the Report

This block consolidates every numerical claim cited in the report's
written sections. Each line corresponds to a specific sentence in
the report. If the underlying data changes, this block reflects it
immediately.

```{r headline-numbers}
cat("===== Headline numbers =====\n\n")

# --- K13A-GRID ---
cat("--- K13A-GRID ---\n")
cat(sprintf("Convergence: GHK %.4f, GQT %.4f\n",
            conv_k13a$converged[conv_k13a$method == "GHK"],
            conv_k13a$converged[conv_k13a$method == "GQT"]))
cat(sprintf("GQT speedup at n=25: %.2fx, n=49: %.2fx\n",
            runtime_wide$ratio_GHK_to_GQT[runtime_wide$n_total == 25],
            runtime_wide$ratio_GHK_to_GQT[runtime_wide$n_total == 49]))
cat(sprintf("Max |MAE %% diff| GHK vs GQT across cells: %.1f%%\n\n",
            max(abs(mae_k13a_wide$pct_diff), na.rm = TRUE)))

# --- MISSPEC ---
cat("--- MISSPEC ---\n")
get_mae <- function(cell, m, p) {
  v <- mae_misspec$MAE[mae_misspec$cell_label == cell &
                       mae_misspec$method == m &
                       mae_misspec$parameter == p]
  if (length(v) == 0) NA else v
}
cat(sprintf("Range MAE NB->NB GHK: %.3f; NB->Poisson GHK: %.3f (%.2fx)\n",
            get_mae("negbin -> negbin", "GHK", "range"),
            get_mae("negbin -> poisson", "GHK", "range"),
            get_mae("negbin -> poisson", "GHK", "range") /
            get_mae("negbin -> negbin", "GHK", "range")))
cat(sprintf("Mean estimated range under NB->Poisson (GHK): %.3f (truth 0.3)\n",
            mean(long_misspec$estimate[long_misspec$cell_label == "negbin -> poisson" &
                                       long_misspec$method == "GHK" &
                                       long_misspec$parameter == "range"])))
cat(sprintf("Mean GQT log-likelihood under NB->Poisson: %.2e\n\n",
            mean(long_misspec$loglik[long_misspec$cell_label == "negbin -> poisson" &
                                     long_misspec$method == "GQT" &
                                     long_misspec$parameter == "Intercept"])))

# --- MISSING ---
cat("--- MISSING ---\n")
for (param in c("Intercept", "range")) {
  for (m in c("GHK", "GQT")) {
    base <- mae_miss$MAE[mae_miss$missing_rate == 0   &
                         mae_miss$parameter == param  &
                         mae_miss$method == m]
    high <- mae_miss$MAE[mae_miss$missing_rate == 0.5 &
                         mae_miss$parameter == param  &
                         mae_miss$method == m]
    cat(sprintf("  %s/%s : 0%% -> 50%%  %.4f -> %.4f  (%.2fx)\n",
                param, m, base, high, high / base))
  }
}
# Bias support: largest |bias| across all rates
cat(sprintf("  Max |bias|: Intercept %.4f; range %.4f (across all rates, both methods)\n",
            max(abs(bias_miss$bias[bias_miss$parameter == "Intercept"])),
            max(abs(bias_miss$bias[bias_miss$parameter == "range"]))))
# Runtime support
cat(sprintf("  Runtime GHK 0%% -> 50%%: %.3f s -> %.3f s\n",
            runtime_miss_wide$runtime_sec.GHK[runtime_miss_wide$missing_rate == 0],
            runtime_miss_wide$runtime_sec.GHK[runtime_miss_wide$missing_rate == 0.5]))
cat("\n")

# --- IRREGULAR ---
cat("--- IRREGULAR ---\n")
get_irr <- function(layout, m, p) {
  v <- mae_irr$MAE[mae_irr$scenario_id == layout &
                   mae_irr$method == m &
                   mae_irr$parameter == p]
  if (length(v) == 0) NA else v
}
cat(sprintf("Intercept MAE grid: %.3f; clustered: %.3f (+%.0f%%)\n",
            get_irr("grid", "GHK", "Intercept"),
            get_irr("clustered", "GHK", "Intercept"),
            100 * (get_irr("clustered", "GHK", "Intercept") /
                   get_irr("grid", "GHK", "Intercept") - 1)))
cat(sprintf("Range MAE grid: %.3f; clustered: %.3f\n\n",
            get_irr("grid", "GHK", "range"),
            get_irr("clustered", "GHK", "range")))

# --- ZIP ---
cat("--- ZIP ---\n")
get_zip <- function(p0v, ff, m, p) {
  v <- mae_zip$MAE[mae_zip$p0 == p0v &
                   mae_zip$fit_family == ff &
                   mae_zip$method == m &
                   mae_zip$parameter == p]
  if (length(v) == 0) NA else v
}
cat(sprintf("Range MAE p0=0.2 ZIP fit GHK: %.3f; Poisson fit GHK: %.3f (+%.0f%%)\n",
            get_zip(0.2, "zip", "GHK", "range"),
            get_zip(0.2, "poisson", "GHK", "range"),
            100 * (get_zip(0.2, "poisson", "GHK", "range") /
                   get_zip(0.2, "zip", "GHK", "range") - 1)))
cat(sprintf("Range MAE p0=0.4 ZIP-GHK: %.3f; ZIP-GQT: %.3f (GQT instability)\n",
            get_zip(0.4, "zip", "GHK", "range"),
            get_zip(0.4, "zip", "GQT", "range")))
# Convergence support for the GQT instability story
cat(sprintf("Min ZIP-GQT convergence: %.4f at p0=%.1f\n",
            min(conv_zip$converged[conv_zip$fit_family == "zip" &
                                   conv_zip$method == "GQT"]),
            conv_zip$p0[conv_zip$fit_family == "zip" &
                        conv_zip$method == "GQT" &
                        conv_zip$converged ==
                          min(conv_zip$converged[conv_zip$fit_family == "zip" &
                                                 conv_zip$method == "GQT"])][1]))
```

# Files Produced

All tables saved as CSV under `tables/report/` and figures saved as PDF
under `visualization/report/`. The naming convention matches the
references in the LaTeX report:

- `tables/report/k13a_runtime.csv`, `tables/report/k13a_mae_wide.csv`,
  `tables/report/k13a_bias.csv`
- `tables/report/misspec_mae.csv`
- `tables/report/missing_mae.csv`, `tables/report/missing_bias.csv`,
  `tables/report/missing_runtime.csv`
- `tables/report/irregular_mae.csv`
- `tables/report/zip_mae_wide.csv`, `tables/report/zip_convergence.csv`
- `visualization/report/k13a_mae_intercept.pdf`, `k13a_mae_range.pdf`
- `visualization/report/misspec_estimates.pdf`
- `visualization/report/missing_mae_vs_rate.pdf`
- `visualization/report/irregular_layouts.pdf`, `irregular_mae.pdf`
- `visualization/report/zip_mae_vs_p0.pdf`

# Session Info

```{r session-info}
sessionInfo()
```
