---
title: 'Missingness Analysis: MCAR Robustness'
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = TRUE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
```

## What this Rmd does

Analyzes the MCAR missingness study: same Poisson(5) baseline as
K13-A default, fitted under five missingness rates: 0%, 10%, 20%,
30%, 50%.

**Question.** How does estimator accuracy degrade as observations
are randomly missing? Does GQT degrade at the same rate as GHK, or
does one of them cope better with incomplete data?

---

## Load and tidy

```{r load}
source("code/R/logging.R")
source("code/R/io.R")
source("code/R/fit_core.R")

DATASET <- "MISSING"

ids <- list_scenarios(DATASET, dir = "data/generated")
fits <- lapply(ids, function(id) load_fit_result(DATASET, id))
names(fits) <- ids

length(fits)   # should be 5
```

### Extract rates from each scenario's config

```{r extract_rate}
# missing_rate lives in spec$missing_rate per the builder. The saved
# scenario config preserves it verbatim.
# We also double-check by measuring the actual NA fraction in the
# scenario file.
rate_info <- do.call(rbind, lapply(fits, function(f) {
  cfg <- f$scenario_config
  data.frame(
    scenario_id    = cfg$scenario_id,
    spec_rate      = cfg$missing_rate,
    stringsAsFactors = FALSE
  )
}))
rate_info
```

### Tidy to long format

```{r tidy}
tidy_one <- function(f) {
  r <- f$results
  cfg <- f$scenario_config

  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!r$converged[i]) {
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id = f$scenario_id,
        missing_rate = cfg$missing_rate,
        replicate   = r$replicate[i],
        method      = r$method[i],
        converged   = FALSE,
        runtime_sec = r$runtime_sec[i],
        n_used      = r$n_used[i],
        parameter   = NA_character_,
        estimate    = NA_real_,
        stringsAsFactors = FALSE
      )
      next
    }
    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next
    for (pname in names(est)) {
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id = f$scenario_id,
        missing_rate = cfg$missing_rate,
        replicate   = r$replicate[i],
        method      = r$method[i],
        converged   = TRUE,
        runtime_sec = r$runtime_sec[i],
        n_used      = r$n_used[i],
        parameter   = pname,
        estimate    = as.numeric(est[[pname]]),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, rows)
}

long <- do.call(rbind, lapply(fits, tidy_one))

# Attach truth (log-lambda for Intercept, range for range).
# Every scenario shares the same generating params, so truth is fixed.
long$truth <- NA_real_
long$truth[long$parameter == "Intercept"] <- log(5)
long$truth[long$parameter == "range"]     <- 0.3
long$abs_err <- abs(long$estimate - long$truth)
long$err     <- long$estimate - long$truth

nrow(long)
head(long)
```

---

## Convergence as rate increases

```{r conv_by_rate}
conv_tab <- aggregate(
  converged ~ missing_rate + method,
  data = unique(long[, c("scenario_id", "replicate", "method",
                         "missing_rate", "converged")]),
  FUN  = function(x) sum(x) / length(x)
)
names(conv_tab)[3] <- "conv_rate"
conv_tab
```

**What to look for:** does convergence break down at high missingness
rates? If 50% MCAR causes more failures than 10%, that's the first
signal that missingness matters.

---

## n_used at each rate

Sanity check that masking produced the expected sample size.

```{r n_used}
n_used_tab <- aggregate(
  n_used ~ missing_rate,
  data = long[long$parameter == "Intercept" & long$converged, ],
  FUN  = function(x) c(mean = mean(x), sd = sd(x),
                       min = min(x), max = max(x))
)
n_used_tab
```

At rate 0 should be n=49. At rate 10 should be ~44. At rate 50 should
be ~25. If these don't match, the masking logic may have a bug.

---

## MAE vs missingness rate

The headline plot: how does estimation accuracy degrade as data is
lost?

```{r mae_table}
mae_tab <- aggregate(
  abs_err ~ missing_rate + method + parameter,
  data = long[long$converged, ],
  FUN  = mean
)
names(mae_tab)[4] <- "MAE"

# Split by parameter for readability
mae_intercept <- mae_tab[mae_tab$parameter == "Intercept", ]
mae_range     <- mae_tab[mae_tab$parameter == "range", ]

cat("MAE for Intercept (log lambda):\n")
print(mae_intercept[order(mae_intercept$missing_rate, mae_intercept$method), ],
      row.names = FALSE, digits = 3)

cat("\nMAE for range:\n")
print(mae_range[order(mae_range$missing_rate, mae_range$method), ],
      row.names = FALSE, digits = 3)
```

### Figure: MAE vs rate, one panel per parameter

```{r plot_mae, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))

for (param in c("Intercept", "range")) {

  d <- mae_tab[mae_tab$parameter == param, ]
  y_max <- max(d$MAE, na.rm = TRUE) * 1.1

  plot(NA, xlim = c(0, 0.5), ylim = c(0, y_max),
       xlab = "MCAR missingness rate",
       ylab = "MAE",
       main = sprintf("MAE vs missingness — %s", param))

  for (m in c("GHK", "GQT")) {
    sub <- d[d$method == m, ]
    sub <- sub[order(sub$missing_rate), ]
    lines(sub$missing_rate, sub$MAE,
          col = ifelse(m == "GHK", "firebrick", "steelblue"),
          lwd = 2)
    points(sub$missing_rate, sub$MAE,
           col = ifelse(m == "GHK", "firebrick", "steelblue"),
           pch = 19, cex = 1.2)
  }
  legend("topleft", legend = c("GHK", "GQT"),
         col = c("firebrick", "steelblue"),
         lwd = 2, pch = 19, bty = "n")
}

par(mfrow = c(1, 1))
```

### Save the figure

```{r save_fig, eval = FALSE}
dir.create("visualization/missing", recursive = TRUE, showWarnings = FALSE)

pdf("visualization/missing/mae_vs_rate.pdf", width = 10, height = 5)
par(mfrow = c(1, 2))
for (param in c("Intercept", "range")) {
  d <- mae_tab[mae_tab$parameter == param, ]
  y_max <- max(d$MAE, na.rm = TRUE) * 1.1
  plot(NA, xlim = c(0, 0.5), ylim = c(0, y_max),
       xlab = "MCAR missingness rate", ylab = "MAE",
       main = sprintf("MAE vs missingness — %s", param))
  for (m in c("GHK", "GQT")) {
    sub <- d[d$method == m, ]
    sub <- sub[order(sub$missing_rate), ]
    lines(sub$missing_rate, sub$MAE,
          col = ifelse(m == "GHK", "firebrick", "steelblue"), lwd = 2)
    points(sub$missing_rate, sub$MAE,
           col = ifelse(m == "GHK", "firebrick", "steelblue"), pch = 19, cex = 1.2)
  }
  legend("topleft", legend = c("GHK", "GQT"),
         col = c("firebrick", "steelblue"), lwd = 2, pch = 19, bty = "n")
}
par(mfrow = c(1, 1))
dev.off()
```

---

## Bias vs rate

Separate from MAE: does missingness introduce systematic error?

```{r bias_table}
bias_tab <- aggregate(
  err ~ missing_rate + method + parameter,
  data = long[long$converged, ],
  FUN  = mean
)
names(bias_tab)[4] <- "bias"

bias_intercept <- bias_tab[bias_tab$parameter == "Intercept", ]
bias_range     <- bias_tab[bias_tab$parameter == "range", ]

cat("Bias for Intercept:\n")
print(bias_intercept, row.names = FALSE, digits = 3)

cat("\nBias for range:\n")
print(bias_range, row.names = FALSE, digits = 3)
```

**What to look for:** under MCAR, estimators should remain
*unbiased* even as variance increases with missingness. If bias
grows with the rate, that's a more serious finding than MAE growth
(because MAE can increase purely from higher variance).

---

## Runtime as rate increases

Does missingness speed things up? (Fewer data points per fit = fewer
likelihood evaluations.)

```{r runtime_by_rate}
rt <- long[long$converged & long$parameter == "Intercept", ]

rt_tab <- aggregate(
  runtime_sec ~ missing_rate + method,
  data = rt,
  FUN  = mean
)
rt_tab
```

---

## Summary

```{r summary}
cat("=== Missingness study summary ===\n\n")

cat("Convergence rates by missingness:\n")
print(conv_tab, row.names = FALSE)

cat("\nRelative MAE increase from 0% to 50% missing:\n")
for (param in c("Intercept", "range")) {
  for (m in c("GHK", "GQT")) {
    base_mae <- mae_tab$MAE[mae_tab$missing_rate == 0 &
                            mae_tab$method == m &
                            mae_tab$parameter == param]
    max_mae <- mae_tab$MAE[mae_tab$missing_rate == 0.5 &
                           mae_tab$method == m &
                           mae_tab$parameter == param]
    if (length(base_mae) == 1 && length(max_mae) == 1) {
      cat(sprintf("  %s / %s: baseline MAE=%.4f  -> 50%% MAE=%.4f  (%.1fx)\n",
                  param, m, base_mae, max_mae, max_mae / base_mae))
    }
  }
}
```

### Likely findings to include in the report

Fill in as numbers come out:

1. **Convergence.** Both methods maintain ___% convergence even at
   50% missingness.
2. **Bias.** Remains near zero across all rates (expected under MCAR).
3. **Accuracy degradation.** MAE grows approximately linearly with
   missingness rate. At 50%, MAE is roughly ___x the baseline.
4. **Method comparison.** GHK and GQT degrade at similar rates — no
   evidence that one method is more robust to MCAR than the other.
5. **Runtime.** Runtime decreases as rate increases, due to smaller
   effective sample size.

---

## End of missingness analysis
