---
title: 'ZIP Analysis: Zero-Inflation and Its Misspecification'
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = TRUE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

## Purpose

Analyzes the ZIP (zero-inflated Poisson) study. Three generating
levels of zero-inflation (p0 = 0, 0.2, 0.4), each fit two ways:

- **Well-specified:** fit as ZIP.
- **Misspecified:** fit as Poisson (ignoring the excess zeros).

Question: does ignoring zero-inflation systematically bias estimates?
Does GHK cope better than GQT, or vice versa?

---

## Load and tidy

```{r load}
source("code/R/logging.R")
source("code/R/io.R")
source("code/R/fit_core.R")

DATASET <- "ZIP"
inventory <- list_fit_results(DATASET, dir = "data/fits")
inventory
```

The inventory should show 6 rows: 3 scenarios × {ZIP fit, Poisson
fit}. The `misspec_family` column is NA for ZIP fits and
`"poisson"` for misspecified fits.

```{r load_all}
fits <- lapply(seq_len(nrow(inventory)), function(i) {
  load_fit_result(
    DATASET,
    inventory$scenario_id[i],
    misspec_family = if (is.na(inventory$misspec_family[i])) NULL else inventory$misspec_family[i]
  )
})
names(fits) <- paste(inventory$scenario_id,
                     ifelse(is.na(inventory$misspec_family),
                            "wellspec", inventory$misspec_family),
                     sep = "__")
names(fits)
```

### Tidy to long

```{r tidy}
tidy_one <- function(f) {
  r   <- f$results
  cfg <- f$scenario_config

  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!r$converged[i]) next
    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next
    for (pname in names(est)) {
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id   = f$scenario_id,
        p0            = cfg$marginal$od,
        fit_family    = f$fit_marginal_spec$family,
        misspecified  = f$misspecified,
        replicate     = r$replicate[i],
        method        = r$method[i],
        runtime_sec   = r$runtime_sec[i],
        parameter     = pname,
        estimate      = as.numeric(est[[pname]]),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, rows)
}

long <- do.call(rbind, lapply(fits, tidy_one))
nrow(long)
head(long)
```

### Attach truth

```{r attach_truth}
long$truth <- NA_real_
long$truth[long$parameter %in% "Intercept"] <- log(5)   # lambda = 5
long$truth[long$parameter %in% "range"]     <- 0.3

long$abs_err <- abs(long$estimate - long$truth)
long$err     <- long$estimate - long$truth
```

---

## Convergence by cell

```{r conv}
conv <- aggregate(
  converged ~ p0 + fit_family + method,
  data = unique(do.call(rbind, lapply(fits, function(f) {
    cbind(p0 = f$scenario_config$marginal$od,
          fit_family = f$fit_marginal_spec$family,
          f$results[, c("replicate", "method", "converged")])
  }))),
  FUN = function(x) mean(x)
)
conv
```

---

## MAE comparison

```{r mae}
mae_tab <- aggregate(
  abs_err ~ p0 + fit_family + method + parameter,
  data = long,
  FUN  = mean
)
names(mae_tab)[5] <- "MAE"

mae_tab[order(mae_tab$parameter, mae_tab$p0, mae_tab$fit_family,
              mae_tab$method), ]
```

### Wide view: well-specified vs misspecified, side by side

```{r mae_wide}
mae_wide <- reshape(
  mae_tab,
  idvar = c("p0", "method", "parameter"),
  timevar = "fit_family",
  direction = "wide"
)
mae_wide$diff <- mae_wide$MAE.poisson - mae_wide$MAE.zip
mae_wide$pct  <- 100 * mae_wide$diff / mae_wide$MAE.zip
mae_wide[order(mae_wide$parameter, mae_wide$p0, mae_wide$method), ]
```

**Report claim shape:** "Fitting Poisson to ZIP data at p0 = 0.4
inflates MAE on the intercept by roughly X%..."

### Figure

```{r plot_mae, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))

for (param in c("Intercept", "range")) {
  d <- mae_tab[mae_tab$parameter == param, ]
  y_max <- max(d$MAE) * 1.1
  plot(NA, xlim = c(0, 0.45), ylim = c(0, y_max),
       xlab = "zero-inflation probability p0",
       ylab = "MAE",
       main = sprintf("MAE vs p0 -- %s", param))
  for (ff in c("zip", "poisson")) {
    for (m in c("GHK", "GQT")) {
      sub <- d[d$fit_family == ff & d$method == m, ]
      sub <- sub[order(sub$p0), ]
      col  <- if (ff == "zip") "darkgreen" else "firebrick"
      lty  <- if (m == "GHK") 1 else 2
      lines(sub$p0, sub$MAE, col = col, lty = lty, lwd = 2)
      points(sub$p0, sub$MAE, col = col, pch = 19, cex = 1.2)
    }
  }
  legend("topleft",
         legend = c("ZIP fit (well-spec)", "Poisson fit (misspec)",
                    "GHK (solid)", "GQT (dashed)"),
         col = c("darkgreen", "firebrick", "black", "black"),
         lty = c(1, 1, 1, 2), lwd = 2, bty = "n", cex = 0.85)
}
par(mfrow = c(1, 1))
```

### Save

```{r save_fig, eval = TRUE}
dir.create("visualization/zip", recursive = TRUE, showWarnings = TRUE)
pdf("visualization/zip/mae_vs_p0.pdf", width = 10, height = 5)
# <paste the plot_mae code here>
dev.off()
```

---

## Summary

```{r summary}
cat("=== ZIP study summary ===\n\n")

for (p0_val in c(0, 0.2, 0.4)) {
  cat(sprintf("\n-- p0 = %g --\n", p0_val))
  for (param in c("Intercept", "range")) {
    zip_mae <- mean(long$abs_err[long$p0 == p0_val &
                                 long$fit_family == "zip" &
                                 long$parameter == param],
                    na.rm = TRUE)
    pois_mae <- mean(long$abs_err[long$p0 == p0_val &
                                  long$fit_family == "poisson" &
                                  long$parameter == param],
                     na.rm = TRUE)
    cat(sprintf("  %-9s  ZIP fit MAE=%.4f  Poisson fit MAE=%.4f  (ratio=%.2fx)\n",
                param, zip_mae, pois_mae, pois_mae / zip_mae))
  }
}
```

### Likely findings for the report

Fill in:

1. At p0 = 0 (no zero-inflation), ZIP and Poisson fits produce
   equivalent MAE (both well-specified).
2. At p0 > 0, Poisson fits ___ the Intercept by ___% more MAE than
   the ZIP fit.
3. Range estimates are ___ (possibly less affected than Intercept,
   since zero-inflation is a marginal property).
4. GHK and GQT degrade ___ (together / differently).
