---
title: "STAT778 Project: Codebase Walkthrough"
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = TRUE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Purpose of this document

A tour of the codebase. Three goals:

1. Explain the two-level naming system (`dataset_id` vs `scenario_id`)
   and how to manage many datasets.
2. Document every field in the saved scenario object.
3. Describe each file in `code/` and what it contributes.

---

## Part 1: Dataset families and scenarios

### The two identifiers

Every simulation in this project carries two labels:

- **`dataset_id`** — the *family*. Identifies which paper / section a
  simulation recipe comes from. Examples: `"H18-B"` (Han and De Oliveira
  2018, Section 4.4), `"K13-A"` (Kazianka 2013, main estimation study).
- **`scenario_id`** — the specific *instance* within that family.
  The paper target is always named `"default"`. Variants have
  descriptive names like `"range_0.5"` or `"od_2"`.

On disk, each saved simulation becomes one file:

```
data/generated/<dataset_id>/<scenario_id>.rds
```

So a typical layout, once we have variants:

```
data/generated/
└── H18-B/
    ├── default.rds
    ├── range_0.5.rds
    ├── range_0.7.rds
    └── od_2.rds
```

All four files are "H18-B datasets". They share the same family recipe
but differ in specific parameter values.

### Listing what's on disk

```{r list_scenarios}
source("code/R/logging.R")
source("code/R/io.R")

available <- list_scenarios("H18-B")
available
```

### Loading a specific scenario

```{r load_one}
s <- load_scenario("H18-B", "default")
```

### Loading all scenarios for a family

```{r load_all, eval = FALSE}
all_h18b <- lapply(list_scenarios("H18-B"), function(id) {
  load_scenario("H18-B", id)
})
names(all_h18b) <- list_scenarios("H18-B")
```

### Generating multiple scenarios in one run

The paper target lives in `H18B_SCENARIOS$default` and should stay
untouched. To add variants, either edit `code/configs/h18b.R` (for
permanent additions) or append to the list at the R prompt (for
one-offs):

```{r append_variants, eval = FALSE}
source("code/R/spec_utils.R")
source("code/configs/h18b.R")

for (r in c(0.1, 0.5, 0.7)) {
  id <- sprintf("range_%g", r)
  H18B_SCENARIOS[[id]] <- h18b_spec(id, list(corr = list(range = r)))
}

# Then re-run the generation script; it picks up every entry in
# H18B_SCENARIOS and saves one .rds per scenario.
source("code/scripts/01_gen_h18b.R")
```

---

## Part 2: What's inside a saved scenario

A loaded scenario is a list with 14 fields. Here's each one.

```{r show_structure}
str(s, max.level = 1)
```

### Identity

```{r field_id}
s$dataset_id
s$scenario_id
```

Convenience duplicates of `s$config$dataset_id` and
`s$config$scenario_id`, pulled out to the top level so generic code
can find them without descending into `$config`.

### `config` — the full input spec, verbatim

This is the most important field for reproducibility. It is the exact
list that was passed to the simulator. Whatever you put in the config
comes back out here — nothing is dropped, nothing is transformed.

```{r field_config}
str(s$config, max.level = 2)
```

Fields inside `config`:

- `dataset_id`, `scenario_id`, `scenario_type` — identity.
- `layout` — how locations were constructed (grid size, bounds).
- `marginal` — the response distribution family and its parameters.
- `corr` — the spatial correlation family and its parameters.
- `sim_n` — number of independent realizations.
- `seed` — the RNG seed used.

Because the full config is stored, you can always reconstruct exactly
how a scenario was generated, no matter how the pipeline evolves later.

### `locations` — the spatial grid

An $n \times 2$ numeric matrix with columns `x` and `y`.

```{r field_locations}
dim(s$locations)
head(s$locations)
```

### `values` — the simulated observations

An $n \times B$ numeric matrix, where $B = $ `sim_n`. For H18-B,
`sim_n = 1`, so `values` has one column.

```{r field_values}
dim(s$values)
head(s$values)
```

If `sim_n > 1`, each column is an independent replicate:

```r
s$values[, 1]   # first replicate
s$values[, 42]  # 42nd replicate
```

### `values_full` — always NULL at generation time

Reserved for future extensions. If a missingness or contamination step is applied later, the pre-modification values go here; `values` then holds the post-modification data.

```{r field_values_full}
s$values_full
```

### `true_params` — generating parameters in flat form

A flat named list for direct comparison against estimated parameters later. Redundant with what's in `$config`, but conveniently flat.

```{r field_true_params}
s$true_params
```

When we fit `mlegc()` to this data in the analysis stage, the
estimates should land close to these values.

### Provenance: `created_at`, `r_version`, `gcKrig_version`

Records when and with what software this scenario was generated. Reproducibility insurance: if you rerun a year from now with a different `gcKrig` version and get different numbers, these tell you why.

```{r field_provenance}
s$created_at
s$r_version
s$gcKrig_version
```

### `seed_info` — how the RNG was handled

```{r field_seed_info}
s$seed_info
```

- `scheme = "direct"` — the seed was applied via `set.seed()` right before `simgc()` was called.
- `scheme = "fixed"` — used by K13-B/C where observations are prescribed, not drawn (no RNG involved).

### `coord_transform_applied`

```{r field_coord_transform}
s$coord_transform_applied
```

`FALSE` for all current scenarios. Reserved for future anisotropy extensions where coordinates are transformed before simulation.

### `missingness_spec`, `contamination_spec`

```{r field_extensions}
s$missingness_spec
s$contamination_spec
```

NULL placeholders for planned extensions (informative missingness, outlier contamination). Kept in the schema now so the object layout stays stable when those features arrive.

---

## Part 3: What each file in `code/` does

### `code/R/` — shared library (dataset-agnostic)

These five files contain all the reusable machinery. They know nothing about any specific dataset.

#### `code/R/logging.R`

Provides timestamped, leveled log messages used throughout the pipeline.

- `log_info(fmt, ...)` — normal progress messages.
- `log_warn(fmt, ...)` — warnings, always shown.
- `log_debug(fmt, ...)` — diagnostic messages, suppressed unless
  `options(sim_pipeline.debug = TRUE)`.
- `log_rule(label)` — horizontal separator for readability.

#### `code/R/grids.R`

Builds spatial location matrices.

- `make_locations(layout_type, ...)` — single entry point. Currently
  only `layout_type = "regular_grid"` is implemented. Later extensions
  will add `"clustered"`, `"sparse"`, `"transect"`, etc. as new
  branches.

#### `code/R/spec_utils.R`

- `merge_spec(base, overrides)` — recursive list merge. Lets you
  override one leaf of a nested list without restating the rest.
  This is what makes `h18b_spec("range_0.5", list(corr = list(range = 0.5)))`
  work: only `corr$range` changes; `corr$kappa`, `corr$nugget`,
  and everything outside `corr` inherit from the paper target.

#### `code/R/sim_core.R`

The heart of the pipeline. Three public functions.

- `make_scenario(config, locations, values, ...)` — scenario object
  factory. Defines the 14-field schema. All other generators funnel
  through this so the schema lives in exactly one place.
- `simulate_mc_scenario(spec, locations, marginal, corr, ...)` —
  wraps `gcKrig::simgc()`. Handles seeding, shape normalization,
  and logging. Returns a scenario object.
- `build_fixed_scenario(spec, locations, ...)` — for K13-B/C, where
  observations are prescribed rather than simulated. Returns the same
  scenario object type.

The statistics are entirely delegated to `gcKrig::simgc()`. This file
does not implement simulation; it only wraps it in reproducibility
scaffolding.

#### `code/R/io.R`

Persistence.

- `save_scenario(scenario, dir)` — writes `scenario` to
  `<dir>/<dataset_id>/<scenario_id>.rds`.
- `load_scenario(dataset_id, scenario_id, dir)` — reads back the
  file at that path. Equivalent to
  `readRDS(file.path(dir, dataset_id, paste0(scenario_id, ".rds")))`.
- `list_scenarios(dataset_id, dir)` — lists scenario ids currently
  saved under a given family.

### `code/configs/` — per-dataset configurations

One file per dataset family.

#### `code/configs/h18b.R`

- `h18b_spec(scenario_id, overrides)` — builder. Called with no
  arguments, returns the paper target. Called with overrides, returns
  a variant.
- `H18B_SCENARIOS` — named list of specs to generate. Keyed by
  `scenario_id`. Starts with just `default` (the paper target);
  variants can be appended.

Every dataset family will have an analogous file (`k13a.R`, etc.) with
its own builder and scenarios list. Convention: the builder is named
`<dataset>_spec()` and the list is `<DATASET>_SCENARIOS`.

### `code/scripts/` — generation scripts

One script per dataset family. Each is a standalone R script that:

1. Sources the shared library.
2. Sources its dataset's config.
3. Iterates over the scenarios list and calls the appropriate
   simulator for each.

#### `code/scripts/01_gen_h18b.R`

- Resolves the project root (via `$GCKRIG_PROJECT_ROOT`,
  `$SLURM_SUBMIT_DIR`, or `getwd()` in that priority order).
- Sources all needed library files.
- Accepts optional command-line arguments to filter which scenarios
  to generate:

```bash
Rscript code/scripts/01_gen_h18b.R                       # all scenarios
Rscript code/scripts/01_gen_h18b.R default               # only one
Rscript code/scripts/01_gen_h18b.R default range_0.5     # two
```

- Contains three small helpers — `build_marginal()`, `build_corr()`,
  `extract_true_params()` — that translate spec lists into `gcKrig` S3
  objects. These are local to the script for now because they may need
  family-specific branches as new datasets are added; if they end up
  shared across all scripts, they'll be promoted to `code/R/`.

### `code/slurm/` — batch submission

#### `code/slurm/_env.sh`

Shared environment setup. Sourced by every `.slurm` script. This is
where R module loads go — you should have edited it during initial
setup to point at your Hopper R module.

#### `code/slurm/01_gen_h18b.slurm`

`sbatch` submission script for H18-B. Requests 1 CPU, 4 GB RAM, 30
minutes. Calls `Rscript code/scripts/01_gen_h18b.R`. Supports optional
scenario filtering via an environment variable:

```bash
sbatch code/slurm/01_gen_h18b.slurm                                        # all
sbatch --export=ALL,SCENARIO_IDS="default range_0.5" code/slurm/01_gen_h18b.slurm
```

---

## Part 4: End-to-end example

Putting all the pieces together: generate the paper target, load it
back, and verify the round trip works.

### Generate

```{r demo_generate}
source("code/scripts/01_gen_h18b.R")
```

### List what's now on disk

```{r demo_list}
list_scenarios("H18-B")
```

### Load and inspect

```{r demo_load}
s <- load_scenario("H18-B", "default")
```

### Confirm reproducibility

The saved config should reconstruct the paper target exactly:

```{r demo_roundtrip}
source("code/R/spec_utils.R")
source("code/configs/h18b.R")

paper_target <- h18b_spec("default")

identical(s$config, paper_target)
```

`TRUE` means the round trip is clean — no silent drops, no silent
transformations.

### Summary of the data

```{r demo_summary}
values <- s$values[, 1]

cat("n locations:", nrow(s$locations), "\n")
cat("n values:   ", length(values), "\n")
cat("range:      [", min(values), ",", max(values), "]\n")
cat("mean:       ", round(mean(values), 2), "\n")
cat("zeros:      ", sum(values == 0), "\n")
```

---

## Adding new datasets later

When we implement K13-A, K13-B, K13-C, the pattern will be the same.
For each new family, create:

1. `code/configs/<name>.R` — builder `<name>_spec()` + `<NAME>_SCENARIOS`.
2. `code/scripts/<NN>_gen_<name>.R` — load library, iterate scenarios, save.
3. `code/slurm/<NN>_gen_<name>.slurm` — submission script.

Nothing in `code/R/` needs to change. The `build_marginal()` and
`build_corr()` helpers inside each script will grow a new branch when
a new marginal or correlation family is needed, but the core library
stays dataset-agnostic.


---


---

## Visualizing a scenario

```{r viz_setup}
source("code/R/logging.R")
source("code/R/io.R")
source("code/R/helper_functions.R")
library(gcKrig)

s      <- load_scenario("H18-B", "default")
values <- s$values[, 1]
locs   <- s$locations
```

### Set up output folder

```{r viz_save_setup}
viz_dir <- "visualization/example"
```

### Option 1: gcKrig's built-in plotter

```{r viz_gckrig, fig.width = 7, fig.height = 6}
sim_like <- list(
  data = s$values[, 1],
  locs = s$locations
)
class(sim_like) <- "simgc"

plot(sim_like, index = 1, plottype = "Text", col = 2)
plot(sim_like, index = 1, plottype = "Dot",  col = 2)
plot(sim_like, index = 1, plottype = "3D",   col = 2)

save_base_plot(
  quote(plot(sim_like, index = 1, plottype = "Text", col = 2)),
  "01_gckrig_text"
)
save_base_plot(
  quote(plot(sim_like, index = 1, plottype = "Dot", col = 2)),
  "01_gckrig_dot"
)
save_base_plot(
  quote(plot(sim_like, index = 1, plottype = "3D", col = 2)),
  "01_gckrig_3d"
)
```

### Option 2: Data in matrix form

```{r viz_df}
df <- data.frame(
  x     = locs[, "x"],
  y     = locs[, "y"],
  count = values
)

head(df)

write.csv(df, file.path(viz_dir, "02_data_long.csv"), row.names = FALSE)
message("Saved: ", file.path(viz_dir, "02_data_long.csv"))
```

```{r viz_matrix}
count_mat <- matrix(values, nrow = 11, ncol = 11, byrow = FALSE)

count_mat

write.csv(count_mat, file.path(viz_dir, "02_data_matrix.csv"), row.names = FALSE)
message("Saved: ", file.path(viz_dir, "02_data_matrix.csv"))
```

### Option 3: Base R image / contour

```{r viz_image, fig.width = 7, fig.height = 6}
x_seq <- sort(unique(locs[, "x"]))
y_seq <- sort(unique(locs[, "y"]))

draw_image <- function() {
  image(
    x = x_seq, y = y_seq, z = count_mat,
    col  = hcl.colors(20, "YlOrRd", rev = TRUE),
    main = "H18-B simulated counts",
    xlab = "x", ylab = "y"
  )
  contour(x_seq, y_seq, count_mat, add = TRUE, col = "gray30")
  text(locs[, "x"], locs[, "y"], labels = values, cex = 0.6)
}

draw_image()

save_base_plot(quote(draw_image()), "03_image_contour")
```

### Option 4: ggplot2

```{r viz_ggplot, fig.width = 7, fig.height = 6}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  p <- ggplot(df, aes(x = x, y = y, fill = count)) +
    geom_tile() +
    geom_text(aes(label = count), size = 3, color = "white") +
    scale_fill_viridis_c(option = "plasma") +
    coord_equal() +
    labs(title = "H18-B simulated counts",
         x = "x coordinate", y = "y coordinate") +
    theme_minimal()

  print(p)

  ggsave(file.path(viz_dir, "04_ggplot.png"), p,
         width = 7, height = 6, dpi = 150)
  ggsave(file.path(viz_dir, "04_ggplot.pdf"), p,
         width = 7, height = 6)
  message("Saved: ", file.path(viz_dir, "04_ggplot.{png,pdf}"))
} else {
  message("Install ggplot2 with install.packages('ggplot2') to use this.")
}
```

### Option 5: Comparing scenarios side by side

```{r viz_compare, fig.width = 12, fig.height = 5}
ids <- list_scenarios("H18-B")

if (length(ids) >= 2) {
  draw_compare <- function() {
    par(mfrow = c(1, length(ids)))
    on.exit(par(mfrow = c(1, 1)), add = TRUE)
    for (id in ids) {
      sc <- load_scenario("H18-B", id)
      v  <- sc$values[, 1]
      plot(
        sc$locations[, "x"], sc$locations[, "y"],
        pch  = 19,
        cex  = 0.5 + 2.5 * v / max(v),
        col  = hcl.colors(50, "YlOrRd", rev = TRUE)[
          cut(v, breaks = 50, labels = FALSE)
        ],
        xlab = "x", ylab = "y",
        main = sprintf("%s (range = %g)", id, sc$config$corr$range)
      )
    }
  }

  draw_compare()

  save_base_plot(quote(draw_compare()), "05_compare",
                 width = 12, height = 5)
} else {
  message("Only one scenario on disk; skipping comparison plot.")
}
```

### Option 6: Interactive plotly

```{r viz_plotly, fig.width = 7, fig.height = 6}
if (requireNamespace("plotly", quietly = TRUE)) {
  library(plotly)

  p_int <- plot_ly(
    df,
    x        = ~x, y = ~y,
    type     = "scatter", mode = "markers",
    marker   = list(
      size       = ~count * 3 + 5,
      color      = ~count,
      colorscale = "Viridis",
      showscale  = TRUE
    ),
    text      = ~paste("count:", count),
    hoverinfo = "text"
  ) %>%
    layout(title = "H18-B counts (hover for values)")

  p_int

  if (requireNamespace("htmlwidgets", quietly = TRUE)) {
    out_html <- file.path(viz_dir, "06_plotly.html")
    htmlwidgets::saveWidget(p_int, out_html, selfcontained = TRUE)
    message("Saved: ", out_html)
  } else {
    message("Install htmlwidgets to save plotly output: install.packages('htmlwidgets')")
  }
} else {
  message("Install plotly with install.packages('plotly') to use this.")
}
```

### What got saved

```{r viz_list}
list.files(viz_dir, full.names = FALSE)
```