---
title: "H18-B: gcKrig Pipeline Example"
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include = FALSE}
# This Rmd lives at the repo root (STAT778_Project/h18b_example.Rmd).
# Every source code file is under code/; outputs go to data/ and logs/.
# Nothing needs to be changed here because the working directory is
# already the repo root.

knitr::opts_chunk$set(
  message = TRUE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 6
)
```

## What this notebook does

Runs the H18-B generation pipeline interactively and inspects the
result. This is the exact same work the Slurm job does, but inside
RStudio where you can see each step.

---

## One-liner: run the full pipeline

```{r run_full_pipeline}
source("code/scripts/01_gen_h18b.R")
```

That's it. One file on disk: `data/generated/H18-B/default.rds`.
Everything below is either inspecting that file or walking through
the pipeline step by step for learning.

---

## Inspect the output

```{r inspect}
source("code/R/logging.R")
source("code/R/io.R")

s <- load_scenario("H18-B", "default")

str(s, max.level = 1)
```

The full input spec is preserved verbatim:

```{r config_roundtrip}
str(s$config, max.level = 2)
```

Summary of the simulated counts:

```{r summary}
values <- s$values[, 1]
summary(values)
cat("zeros:", sum(values == 0), "of", length(values), "\n")
```

---

## Visualize

```{r plot}
plot(
  s$locations[, 1], s$locations[, 2],
  pch = 19,
  cex = 1 + 2 * values / max(values),
  col = gray(1 - values / max(values)),
  xlab = "x", ylab = "y",
  main = "H18-B simulated counts (darker/bigger = higher)"
)
```

---

## Optional: step through the pipeline manually

If you want to see what happens at each stage:

```{r step_by_step, eval = FALSE}
library(gcKrig)

source("code/R/logging.R")
source("code/R/grids.R")
source("code/R/spec_utils.R")
source("code/R/sim_core.R")
source("code/R/io.R")
source("code/configs/h18b.R")

# Pick the paper target scenario
spec <- H18B_SCENARIOS$default

# 1. Build locations from the layout spec
locs <- do.call(make_locations, spec$layout)

# 2. Build gcKrig objects from the marginal + corr specs
mu_vec <- exp(spec$marginal$beta[1] +
              spec$marginal$beta[2] * locs[, 1] +
              spec$marginal$beta[3] * locs[, 2])

marginal_obj <- negbin.gc(mu = mu_vec, od = spec$marginal$od)
corr_obj     <- matern.gc(range  = spec$corr$range,
                          kappa  = spec$corr$kappa,
                          nugget = spec$corr$nugget)

# 3. Simulate (the wrapper handles seeding and logging)
scenario <- simulate_mc_scenario(
  spec      = spec,
  locations = locs,
  marginal  = marginal_obj,
  corr      = corr_obj
)

# 4. Save
save_scenario(scenario)
```

---

## Optional: try a variant without editing any file

```{r variant, eval = FALSE}
library(gcKrig)
source("code/R/logging.R")
source("code/R/grids.R")
source("code/R/spec_utils.R")
source("code/R/sim_core.R")
source("code/R/io.R")
source("code/configs/h18b.R")

variant <- h18b_spec("range_0.7", list(corr = list(range = 0.7)))

locs_v <- do.call(make_locations, variant$layout)
mu_v   <- exp(variant$marginal$beta[1] +
              variant$marginal$beta[2] * locs_v[, 1] +
              variant$marginal$beta[3] * locs_v[, 2])

scenario_v <- simulate_mc_scenario(
  spec      = variant,
  locations = locs_v,
  marginal  = negbin.gc(mu = mu_v, od = variant$marginal$od),
  corr      = matern.gc(range  = variant$corr$range,
                        kappa  = variant$corr$kappa,
                        nugget = variant$corr$nugget)
)

save_scenario(scenario_v)

list_scenarios("H18-B")  # should now show "default" and "range_0.7"
```

---

## What's next

- `sbatch code/slurm/01_gen_h18b.slurm` runs the same thing as a batch
  job on a Hopper compute node.
- To add a permanent variant, edit `code/configs/h18b.R` and append
  to `H18B_SCENARIOS` there.
