---
title: 'Irregular Sampling Analysis: Layout Effects on Estimation'
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = TRUE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

## Purpose

Compares estimation accuracy under three spatial layouts:

- **Grid:** regular 7x7 (K13-A default).
- **Random:** 49 uniform points in [0,1]^2.
- **Clustered:** 49 points in 4 Gaussian clusters.

All share the same generating parameters (Poisson lambda=5,
range=0.3). Any difference in estimation accuracy therefore reflects
the layout itself.

---

## Load and tidy

```{r load}
source("code/R/logging.R")
source("code/R/io.R")
source("code/R/fit_core.R")

DATASET <- "IRREGULAR"
ids <- list_scenarios(DATASET, dir = "data/generated")
fits <- lapply(ids, function(id) load_fit_result(DATASET, id))
names(fits) <- ids
length(fits)
```

### Visualize the three layouts

Before analysis, confirm the layouts look as intended.

```{r plot_layouts, fig.width = 12, fig.height = 4}
par(mfrow = c(1, length(fits)))

for (id in names(fits)) {
  locs <- fits[[id]]$scenario_config$layout
  scen <- load_scenario(DATASET, id, dir = "data/generated")
  plot(
    scen$locations[, "x"], scen$locations[, "y"],
    pch = 19, cex = 1.2, col = "steelblue",
    xlab = "x", ylab = "y",
    xlim = c(0, 1), ylim = c(0, 1),
    main = sprintf("%s (n=%d)", id, nrow(scen$locations))
  )
}
par(mfrow = c(1, 1))
```

### Tidy

```{r tidy}
tidy_one <- function(f) {
  r <- f$results
  cfg <- f$scenario_config

  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!r$converged[i]) next
    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next
    for (pname in names(est)) {
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id = f$scenario_id,
        layout_type = cfg$layout$layout_type,
        replicate   = r$replicate[i],
        method      = r$method[i],
        runtime_sec = r$runtime_sec[i],
        parameter   = pname,
        estimate    = as.numeric(est[[pname]]),
        stringsAsFactors = FALSE
      )
    }
  }
  do.call(rbind, rows)
}

long <- do.call(rbind, lapply(fits, tidy_one))

long$truth <- NA_real_
long$truth[long$parameter %in% "Intercept"] <- log(5)
long$truth[long$parameter %in% "range"]     <- 0.3

long$abs_err <- abs(long$estimate - long$truth)
long$err     <- long$estimate - long$truth
```

---

## MAE by layout

```{r mae}
mae_tab <- aggregate(
  abs_err ~ scenario_id + method + parameter,
  data = long,
  FUN  = mean
)
names(mae_tab)[4] <- "MAE"
mae_tab[order(mae_tab$parameter, mae_tab$scenario_id, mae_tab$method), ]
```

### Figure

```{r plot_mae, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))

for (param in c("Intercept", "range")) {
  d <- mae_tab[mae_tab$parameter == param, ]
  layouts <- c("grid", "random", "clustered")
  d$scenario_id <- factor(d$scenario_id, levels = layouts)

  wide <- reshape(d[, c("scenario_id", "method", "MAE")],
                  idvar = "scenario_id", timevar = "method",
                  direction = "wide")

  mat <- rbind(wide$MAE.GHK, wide$MAE.GQT)
  colnames(mat) <- wide$scenario_id
  rownames(mat) <- c("GHK", "GQT")

  barplot(
    mat, beside = TRUE,
    col = c("firebrick", "steelblue"),
    ylab = "MAE", main = sprintf("MAE by layout -- %s", param),
    legend.text = rownames(mat),
    args.legend = list(x = "topright", bty = "n")
  )
}
par(mfrow = c(1, 1))
```

---

## Bias by layout

```{r bias}
bias_tab <- aggregate(
  err ~ scenario_id + method + parameter,
  data = long,
  FUN  = mean
)
names(bias_tab)[4] <- "bias"
bias_tab
```

Regular grid baseline: bias should be near zero for both parameters.
Random layout: expected similar. Clustered: might show more bias
(especially in range) because clustered point patterns have a
different distribution of pairwise distances.

---

## Summary

```{r summary}
cat("=== Irregular sampling summary ===\n\n")

for (layout in c("grid", "random", "clustered")) {
  cat(sprintf("\n-- %s --\n", layout))
  for (param in c("Intercept", "range")) {
    for (m in c("GHK", "GQT")) {
      mae_val <- mean(long$abs_err[long$scenario_id == layout &
                                   long$parameter == param &
                                   long$method == m], na.rm = TRUE)
      cat(sprintf("  %-9s  %s  MAE=%.4f\n", param, m, mae_val))
    }
  }
}
```

### Likely findings

Fill in:

1. All three layouts maintain ___% convergence.
2. Random layout produces MAE ___ (similar to / worse than) grid.
3. Clustered layout ___ (the expected issue is worse range
   identification, since clustered points have fewer large
   distances to probe the correlation at its tail).
