---
title: 'K13A-GRID Analysis: GHK vs GQT Across the Parameter Grid'
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = TRUE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

## What this Rmd does

Analyzes the 18-cell grid study (Day 4): 2 sample sizes x 3 ranges x
3 lambda values, each with 200 replicates fit by both GHK and GQT.
Produces the tables and figures that anchor the report's simulation
results section.

**Organization.** Section 1 tidies the raw fit results into a long
data frame. Sections 2-5 produce specific outputs (convergence,
runtime, estimates, MAE). Section 6 summarizes findings.

---

## Section 1: Load and tidy

### Load every cell

```{r load}
source("code/R/logging.R")
source("code/R/io.R")
source("code/R/fit_core.R")

DATASET <- "K13A-GRID"

ids <- list_scenarios(DATASET, dir = "data/generated")
length(ids)   # should be 18
```

```{r load_all}
# Load all fit results. Each one is a list with dataset_id,
# scenario_id, results (data frame), and metadata.
fits <- lapply(ids, function(id) load_fit_result(DATASET, id))
names(fits) <- ids
```

Quick sanity check — same shape everywhere?

```{r sanity}
sapply(fits, function(f) nrow(f$results))   # should all be 400 (200 x 2)
```

### Extract cell parameters from each scenario_id

The scenario_id encodes (n, range, lambda) so we don't need to dig
into each config. But parsing out of the filename is a fragile
convention — better to pull from the stored config.

```{r extract_cell_params}
cell_info <- do.call(rbind, lapply(fits, function(f) {
  cfg <- f$scenario_config
  data.frame(
    scenario_id = cfg$scenario_id,
    n_total     = cfg$layout$n_side^2,
    range       = cfg$corr$range,
    lambda      = cfg$marginal$lambda,
    stringsAsFactors = FALSE
  )
}))

head(cell_info)
```

### Tidy to long format

One row per (scenario, replicate, method, parameter). This is the
shape that makes aggregation easy.

```{r tidy_long}
tidy_one <- function(f) {
  r <- f$results
  cfg <- f$scenario_config

  rows <- list()
  for (i in seq_len(nrow(r))) {
    if (!r$converged[i]) {
      # record a single "failed" row so failures are visible
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id = f$scenario_id,
        n_total     = cfg$layout$n_side^2,
        range       = cfg$corr$range,
        lambda      = cfg$marginal$lambda,
        replicate   = r$replicate[i],
        method      = r$method[i],
        converged   = FALSE,
        runtime_sec = r$runtime_sec[i],
        parameter   = NA_character_,
        estimate    = NA_real_,
        stringsAsFactors = FALSE
      )
      next
    }

    est <- r$estimates[[i]]
    if (is.null(est) || length(est) == 0) next

    for (pname in names(est)) {
      rows[[length(rows) + 1]] <- data.frame(
        scenario_id = f$scenario_id,
        n_total     = cfg$layout$n_side^2,
        range       = cfg$corr$range,
        lambda      = cfg$marginal$lambda,
        replicate   = r$replicate[i],
        method      = r$method[i],
        converged   = TRUE,
        runtime_sec = r$runtime_sec[i],
        parameter   = pname,
        estimate    = as.numeric(est[[pname]]),
        stringsAsFactors = FALSE
      )
    }
  }

  do.call(rbind, rows)
}

long <- do.call(rbind, lapply(fits, tidy_one))
nrow(long)
head(long)
```

### Attach truth

```{r attach_truth}
# gcKrig's Poisson intercept is on the log scale; truth is log(lambda).
long$truth <- NA_real_
long$truth[long$parameter %in% "Intercept"] <- log(long$lambda[long$parameter %in% "Intercept"])
long$truth[long$parameter %in% "range"]     <- long$range[long$parameter %in% "range"]

long$abs_err <- abs(long$estimate - long$truth)
long$err     <- long$estimate - long$truth   # signed, for bias

head(long[, c("scenario_id", "method", "parameter", "estimate", "truth", "err")])
```

---

## Section 2: Convergence diagnostics

```{r convergence_table}
conv_tab <- aggregate(
  converged ~ n_total + range + lambda + method,
  data = unique(long[, c("scenario_id", "replicate", "method",
                         "n_total", "range", "lambda", "converged")]),
  FUN  = function(x) sum(x) / length(x)
)
names(conv_tab)[5] <- "conv_rate"

# Flag any cell below 99% convergence
conv_tab[conv_tab$conv_rate < 0.99, ]
```

If that's empty, every cell had at least 99% convergence. Good.

```{r convergence_summary}
cat("Overall convergence by method:\n")
print(aggregate(conv_rate ~ method, data = conv_tab, FUN = mean))
```

---

## Section 3: Runtime comparison

```{r runtime_table}
rt <- long[long$converged, ]
rt <- unique(rt[, c("scenario_id", "replicate", "method", "n_total",
                    "range", "lambda", "runtime_sec")])

runtime_summary <- aggregate(
  runtime_sec ~ n_total + method,
  data = rt,
  FUN  = function(x) c(mean = mean(x), median = median(x),
                       min = min(x), max = max(x))
)
runtime_summary
```

### Headline: GHK vs GQT speed ratio

```{r runtime_ratio}
rt_means <- aggregate(runtime_sec ~ n_total + method, data = rt, FUN = mean)
rt_wide <- reshape(rt_means, idvar = "n_total", timevar = "method",
                   direction = "wide")
rt_wide$ratio_GHK_to_GQT <- rt_wide$runtime_sec.GHK / rt_wide$runtime_sec.GQT
rt_wide
```

This is the number for the report: "GQT is roughly Xx faster than GHK
at n=25, and Yx faster at n=49."

### Plot: runtime by cell

```{r plot_runtime, fig.width = 9, fig.height = 5}
par(mfrow = c(1, 2))

for (n_val in sort(unique(rt$n_total))) {
  sub <- rt[rt$n_total == n_val, ]
  boxplot(
    runtime_sec ~ method, data = sub,
    log = "y",
    main = sprintf("Runtime at n=%d (log scale)", n_val),
    ylab = "seconds per fit"
  )
}

par(mfrow = c(1, 1))
```

---

## Section 4: Parameter estimates

### Means and SDs per cell

```{r estimate_summary}
est_summary <- aggregate(
  estimate ~ n_total + range + lambda + method + parameter,
  data = long[long$converged, ],
  FUN  = function(x) c(mean = mean(x), sd = sd(x), n = length(x))
)

# flatten the aggregated value column for display
est_display <- do.call(data.frame, est_summary)
head(est_display)
```

### Bias (signed error)

```{r bias_table}
bias_tab <- aggregate(
  err ~ n_total + range + lambda + method + parameter,
  data = long[long$converged, ],
  FUN  = mean
)
names(bias_tab)[6] <- "bias"

# Split by parameter for readability
bias_intercept <- bias_tab[bias_tab$parameter == "Intercept", ]
bias_range     <- bias_tab[bias_tab$parameter == "range", ]

cat("Bias for log-Intercept:\n")
print(bias_intercept, row.names = FALSE)

cat("\nBias for range:\n")
print(bias_range, row.names = FALSE)
```

**What to look for.** Values close to zero mean the estimate is on
average correct. Range estimates tend to be *negatively biased* —
finite samples underestimate spatial dependence. Intercept estimates
are usually closer to unbiased.

---

## Section 5: MAE — the headline table and figure

### MAE per cell

```{r mae_table}
mae_tab <- aggregate(
  abs_err ~ n_total + range + lambda + method + parameter,
  data = long[long$converged, ],
  FUN  = mean
)
names(mae_tab)[6] <- "MAE"

# Two sub-tables for the report: one per parameter.
mae_intercept <- mae_tab[mae_tab$parameter == "Intercept", ]
mae_range     <- mae_tab[mae_tab$parameter == "range", ]

cat("MAE for log-Intercept:\n")
print(mae_intercept[order(mae_intercept$n_total, mae_intercept$range,
                          mae_intercept$lambda, mae_intercept$method), ],
      row.names = FALSE, digits = 3)

cat("\nMAE for range:\n")
print(mae_range[order(mae_range$n_total, mae_range$range,
                      mae_range$lambda, mae_range$method), ],
      row.names = FALSE, digits = 3)
```

### Direct GHK vs GQT MAE comparison

The cleanest way to see whether GQT keeps up with GHK everywhere is a
wide table: one row per cell, separate columns for each method, and a
difference.

```{r mae_compare_wide}
mae_wide <- reshape(
  mae_tab,
  idvar     = c("n_total", "range", "lambda", "parameter"),
  timevar   = "method",
  direction = "wide"
)
mae_wide$diff <- mae_wide$MAE.GQT - mae_wide$MAE.GHK
mae_wide$pct_diff <- 100 * mae_wide$diff / mae_wide$MAE.GHK

mae_wide <- mae_wide[order(mae_wide$parameter, mae_wide$n_total,
                           mae_wide$range, mae_wide$lambda), ]
print(mae_wide, row.names = FALSE, digits = 3)
```

**Report claim would be something like:** "Across the 18 cells, GQT's
MAE differs from GHK's by less than X% for the intercept and less
than Y% for the range parameter."

### The Figure-1 equivalent: MAE vs sample size

Kazianka's paper plots MAE as a function of n, faceted by range,
with colored lines for different methods or lambda values. Even with
only n in {25, 49}, the plot is informative.

```{r plot_mae_figure, fig.width = 10, fig.height = 6}
# Separate panels for the two parameters. Base-R approach because
# the report may ultimately use ggplot for final; this is the
# exploratory version.

plot_mae_panel <- function(param_name) {
  dat <- mae_tab[mae_tab$parameter == param_name, ]
  ranges <- sort(unique(dat$range))

  par(mfrow = c(1, length(ranges)),
      oma = c(0, 0, 3, 0), mar = c(4, 4, 3, 1))

  for (r in ranges) {
    d <- dat[dat$range == r, ]
    y_max <- max(d$MAE, na.rm = TRUE) * 1.1

    # One line per method; x = n_total; color by lambda
    plot(NA, xlim = range(d$n_total), ylim = c(0, y_max),
         xlab = "n", ylab = "MAE",
         main = sprintf("range = %g", r))

    lams <- sort(unique(d$lambda))
    cols <- c("firebrick", "steelblue", "darkgreen")
    ltys <- c(1, 2)  # solid = GHK, dashed = GQT

    for (li in seq_along(lams)) {
      for (mi in seq_along(c("GHK", "GQT"))) {
        m <- c("GHK", "GQT")[mi]
        sub <- d[d$lambda == lams[li] & d$method == m, ]
        sub <- sub[order(sub$n_total), ]
        lines(sub$n_total, sub$MAE,
              col = cols[li], lty = ltys[mi], lwd = 2)
        points(sub$n_total, sub$MAE,
               col = cols[li], pch = 19)
      }
    }
  }

  mtext(sprintf("MAE vs n for parameter %s  (solid=GHK, dashed=GQT, colors=lambda)",
                param_name),
        outer = TRUE, cex = 1.2)
  par(mfrow = c(1, 1))
}

plot_mae_panel("Intercept")
```

```{r plot_mae_range, fig.width = 10, fig.height = 6}
plot_mae_panel("range")
```

### Save the figures

```{r save_figures, eval = FALSE}
dir.create("visualization/k13a_grid", recursive = TRUE, showWarnings = FALSE)

pdf("visualization/k13a_grid/mae_intercept.pdf", width = 10, height = 6)
plot_mae_panel("Intercept")
dev.off()

pdf("visualization/k13a_grid/mae_range.pdf", width = 10, height = 6)
plot_mae_panel("range")
dev.off()

message("Figures saved to visualization/k13a_grid/")
```

---

## Section 6: Summary for the report

Brief checklist of what the results say. Fill in the actual numbers
after running the chunks above.

```{r summary_stats}
# Convergence
cat("Overall convergence:\n")
print(round(tapply(long$converged[long$parameter == "Intercept" | is.na(long$parameter)],
                   long$method[long$parameter == "Intercept" | is.na(long$parameter)],
                   mean), 4))

# Runtime speedup
cat("\nMean runtime:\n")
print(tapply(rt$runtime_sec, rt$method, mean))
cat(sprintf("\nGQT speedup over GHK: %.1fx\n",
            mean(rt$runtime_sec[rt$method == "GHK"]) /
            mean(rt$runtime_sec[rt$method == "GQT"])))

# Max MAE difference
max_diff_pct <- max(abs(mae_wide$pct_diff), na.rm = TRUE)
cat(sprintf("\nMax |MAE difference| between GHK and GQT: %.2f%%\n", max_diff_pct))
```

### Likely findings to report

Fill in as the numbers come out:

1. **Convergence.** Both methods converge at ___% across the 18
   cells.
2. **Runtime.** GQT is ~___x faster than GHK on average.
3. **Accuracy.** Across all 18 cells, GHK and GQT produce MAE values
   within ___% of each other for both parameters.
4. **Sample size effect.** MAE decreases by roughly a factor of
   ___ going from n=25 to n=49, consistent with a sqrt(n)
   convergence rate.
5. **Range effect.** MAE of the range parameter grows with the true
   range value — larger ranges are harder to identify at fixed n.
6. **Lambda effect.** Lower-intensity cells (lambda=1) have [more/less]
   variability in estimates.
7. **Any surprises.** [Fill in any cell where GQT materially
   underperformed GHK, if any.]

---

## Section 7: Tables for the report

LaTeX-ready tables. These can be pasted into the report directly
(wrapped in a `\begin{tabular}` environment).

```{r latex_runtime}
cat("\n% ---- Runtime table ----\n")
rt_summary_simple <- aggregate(
  runtime_sec ~ n_total + method,
  data = rt,
  FUN  = mean
)

for (i in seq_len(nrow(rt_summary_simple))) {
  cat(sprintf("%d & %s & %.3f \\\\\n",
              rt_summary_simple$n_total[i],
              rt_summary_simple$method[i],
              rt_summary_simple$runtime_sec[i]))
}
```

```{r latex_mae}
cat("\n% ---- MAE table (Intercept) ----\n")
for (i in seq_len(nrow(mae_intercept))) {
  cat(sprintf("%d & %g & %g & %s & %.4f \\\\\n",
              mae_intercept$n_total[i],
              mae_intercept$range[i],
              mae_intercept$lambda[i],
              mae_intercept$method[i],
              mae_intercept$MAE[i]))
}
```

---

## End of K13A-GRID analysis

Next steps:
- `k13a_pilot_analysis.Rmd` — Day-1 pilot (sim_n=200) as a single
  reference point, plus H18-B (sim_n=1) just for the figure.
- `misspec_analysis.Rmd` — 2x2 factorial analysis of the MISSPEC
  study.
- Then the missingness and zero-inflation extensions.
