---
title: "Day 3 Misspecification Analysis"
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Purpose

This document analyzes the **Day 3 misspecification study**.

The current Day 3 design is a (2 $\times$ 2) matrix:

| Generating family | Fitting family | Meaning        |
| ----------------- | -------------- | -------------- |
| Poisson           | Poisson        | well-specified |
| Poisson           | NegBin         | misspecified   |
| NegBin            | NegBin         | well-specified |
| NegBin            | Poisson        | misspecified   |

The goals of this analysis are:

1. verify that the misspecification runs completed successfully,
2. summarize convergence and runtime,
3. extract fitted parameter estimates,
4. compare estimates to the correct truth where that comparison is well-defined,
5. produce the first Day 3 tables and plots for the report.

---

# Background: what the pipeline is doing

From the code standpoint, the Day 3 pipeline has two stages.

## Generation stage

The generation script creates **scenario objects**. For the current misspecification pilot, there are two generating scenarios:

* `pois_pilot`
* `nbin_pilot`

Each scenario contains:

* `locations`: spatial coordinates
* `values`: a matrix of simulated datasets
* `config`: the full generating specification
* `true_params`: the flat generation truth

In both cases:

* the layout is a regular (7 $\times$ 7) grid, so (n = 49),
* the number of Monte Carlo replicates is (B = 30),
* the dependence structure is Matérn with:

  * range (= 0.3),
  * ($\kappa$ = 0.5),
  * nugget (= 0).

Because ($\kappa$ = 0.5), the Matérn reduces to the exponential correlation model
$$
\rho(h) = \exp(-h / r),
$$
with (r = 0.3).

The marginal families differ:

* `pois_pilot`: Poisson with mean ($\lambda$ = 5),
* `nbin_pilot`: negative binomial with constant mean (5) and overdispersion `od = 1`.

So each replicate is a length-49 vector of spatially correlated counts.

## Fitting stage

The fitting script loads a generated scenario and then fits each replicate with:

* `GHK`
* `GQT`

For Day 3, the fitting wrapper can override the marginal family used in fitting. That is how misspecification is implemented.

The key point is:

* the **generating family** comes from `scenario$config$marginal`,
* the **fitting family** comes from `fit_result$fit_marginal_spec`.

The fitting wrapper returns a **fit-result object** with one row per `(replicate, method)` pair.

So if one scenario has:

* 30 replicates
* 2 methods

then the resulting `fit_result$results` table has:

$$
30 \times 2 = 60
$$

rows.

---

# Load packages and project code

```{r load_pipeline}
suppressPackageStartupMessages({
  library(gcKrig)
})

source("code/R/logging.R")
source("code/R/io.R")
source("code/R/fit_core.R")
```

---

# Load all Day 3 fit results

The Day 3 fitting script is expected to write fit-result objects under:

```{}
data/fits/MISSPEC/
```

We load every `.rds` file from that directory.

```{r load_fit_files}
fit_dir <- file.path("data", "fits", "MISSPEC")
fit_files <- list.files(fit_dir, pattern = "\\.rds$", full.names = TRUE)

fit_files
```

```{r read_fit_results}
if (length(fit_files) == 0L) {
  stop("No Day 3 fit-result files found under data/fits/MISSPEC/.")
}

fit_objects <- lapply(fit_files, readRDS)
names(fit_objects) <- basename(fit_files)

names(fit_objects)
```

---

# Inspect one fit-result object

This helps remind us of the structure before building summaries.

```{r inspect_one_fit_result}
example_name <- names(fit_objects)[1]
example_fit  <- fit_objects[[1]]

cat("Example file:", example_name, "\n\n")
str(example_fit, max.level = 1)
```

---

# Helper functions

These helpers flatten the nested fit-result objects into analysis-ready tables.

```{r helpers}
safe_first <- function(x, default = NA) {
  if (length(x) == 0L || is.null(x)) return(default)
  x[[1]]
}

pretty_cell_label <- function(gen_family, fit_family) {
  paste0(gen_family, " \u2192 ", fit_family)
}

flatten_fit_result <- function(fr, source_file = NA_character_) {
  res <- fr$results

  out <- data.frame(
    dataset_id      = fr$dataset_id,
    gen_scenario_id = fr$scenario_id,
    source_file     = source_file,
    gen_family      = fr$scenario_config$marginal$family,
    fit_family      = fr$fit_marginal_spec$family,
    misspecified    = fr$misspecified,
    replicate       = res$replicate,
    method          = res$method,
    converged       = res$converged,
    error_message   = ifelse(is.na(res$error_message), "", res$error_message),
    runtime_sec     = res$runtime_sec,
    loglik          = res$loglik,
    aic             = res$aic,
    bic             = res$bic,
    aicc            = res$aicc,
    stringsAsFactors = FALSE
  )

  out$cell_label <- pretty_cell_label(out$gen_family, out$fit_family)

  # Preserve list-columns
  out$estimates  <- res$estimates
  out$std_errors <- res$std_errors

  out
}

normalize_param_name <- function(pname) {
  p <- tolower(pname)
  p <- gsub("[[:space:]]+", "", p)
  p <- gsub("[`()]", "", p)
  p
}

truth_for_parameter <- function(param_name, gen_config, fit_marginal_spec) {
  p <- normalize_param_name(param_name)

  gen_marg <- gen_config$marginal
  gen_corr <- gen_config$corr
  gen_family <- gen_marg$family
  fit_family <- fit_marginal_spec$family

  # range truth
  if (grepl("range", p)) {
    return(gen_corr$range)
  }

  # kappa truth, only if ever estimated
  if (grepl("kappa", p)) {
    return(gen_corr$kappa)
  }

  # nugget truth, only if ever estimated
  if (grepl("nugget", p)) {
    return(gen_corr$nugget)
  }

  # overdispersion / od truth
  if (grepl("^od$|overdisp|dispersion|disp", p)) {
    if (gen_family == "negbin" && fit_family == "negbin") {
      return(gen_marg$od)
    } else {
      return(NA_real_)
    }
  }

  # intercept / beta0 / mean-scale regression coefficient
  if (grepl("intercept", p) || grepl("^beta0$", p) || grepl("^beta1$", p) ||
      grepl("^beta$", p) || grepl("^mu$", p)) {
    if (gen_family == "poisson") {
      return(log(gen_marg$lambda))
    }
    if (gen_family == "negbin") {
      return(gen_marg$beta[1])
    }
  }

  # fallback: if the generating model has only one marginal coefficient
  # and the parameter isn't a known corr/dispersion parameter, treat it
  # as the intercept-scale mean parameter
  if (gen_family == "poisson") {
    return(log(gen_marg$lambda))
  }
  if (gen_family == "negbin") {
    return(gen_marg$beta[1])
  }

  NA_real_
}

extract_estimates_long <- function(flat_df, fit_objects_named) {
  rows <- list()
  idx  <- 0L

  for (i in seq_len(nrow(flat_df))) {
    if (!isTRUE(flat_df$converged[i])) next

    est <- flat_df$estimates[[i]]
    if (is.null(est) || length(est) == 0L) next

    src_name <- flat_df$source_file[i]
    fr <- fit_objects_named[[src_name]]
    gen_config <- fr$scenario_config
    fit_spec   <- fr$fit_marginal_spec

    for (pname in names(est)) {
      idx <- idx + 1L
      truth_val <- truth_for_parameter(pname, gen_config, fit_spec)

      rows[[idx]] <- data.frame(
        dataset_id      = flat_df$dataset_id[i],
        gen_scenario_id = flat_df$gen_scenario_id[i],
        source_file     = flat_df$source_file[i],
        gen_family      = flat_df$gen_family[i],
        fit_family      = flat_df$fit_family[i],
        misspecified    = flat_df$misspecified[i],
        cell_label      = flat_df$cell_label[i],
        replicate       = flat_df$replicate[i],
        method          = flat_df$method[i],
        parameter       = pname,
        estimate        = as.numeric(est[[pname]]),
        truth           = truth_val,
        abs_error       = if (is.na(truth_val)) NA_real_ else abs(as.numeric(est[[pname]]) - truth_val),
        stringsAsFactors = FALSE
      )
    }
  }

  if (length(rows) == 0L) {
    return(data.frame())
  }

  do.call(rbind, rows)
}
```

---

# Build the master fit table

```{r build_master_table}
fit_df_list <- Map(
  function(obj, nm) flatten_fit_result(obj, source_file = nm),
  fit_objects,
  names(fit_objects)
)

fit_df <- do.call(rbind, fit_df_list)

dim(fit_df)
head(fit_df)
```

---

# Overview of the loaded Day 3 cells

```{r cell_overview}
cell_overview <- unique(
  fit_df[, c("gen_scenario_id", "gen_family", "fit_family", "misspecified", "cell_label", "source_file")]
)

rownames(cell_overview) <- NULL
print(cell_overview)
```

This should show the four expected cells:

* Poisson ($\rightarrow$) Poisson
* Poisson ($\rightarrow$) NegBin
* NegBin ($\rightarrow$) NegBin
* NegBin ($\rightarrow$) Poisson

---

# Convergence summary

We summarize convergence separately by:

* generating family,
* fitting family,
* estimation method.

```{r convergence_summary}
convergence_tab <- aggregate(
  converged ~ gen_family + fit_family + cell_label + method,
  data = fit_df,
  FUN = function(x) c(
    n_total = length(x),
    n_ok    = sum(x),
    rate    = mean(x)
  )
)

convergence_summary <- data.frame(
  gen_family = convergence_tab$gen_family,
  fit_family = convergence_tab$fit_family,
  cell_label = convergence_tab$cell_label,
  method     = convergence_tab$method,
  n_total    = convergence_tab$converged[, "n_total"],
  n_ok       = convergence_tab$converged[, "n_ok"],
  conv_rate  = convergence_tab$converged[, "rate"]
)

print(convergence_summary, digits = 3)
```

Interpretation:

* A low convergence rate means that fitting is unstable for that cell.
* If misspecified cells converge much less often, that is already a meaningful result.

---

# Runtime summary

We now summarize runtime among the converged fits.

```{r runtime_summary}
runtime_df <- fit_df[fit_df$converged, ]

runtime_tab <- aggregate(
  runtime_sec ~ gen_family + fit_family + cell_label + method,
  data = runtime_df,
  FUN = function(x) c(
    mean   = mean(x),
    median = stats::median(x),
    min    = min(x),
    max    = max(x),
    total  = sum(x)
  )
)

runtime_summary <- data.frame(
  gen_family = runtime_tab$gen_family,
  fit_family = runtime_tab$fit_family,
  cell_label = runtime_tab$cell_label,
  method     = runtime_tab$method,
  mean_sec   = runtime_tab$runtime_sec[, "mean"],
  median_sec = runtime_tab$runtime_sec[, "median"],
  min_sec    = runtime_tab$runtime_sec[, "min"],
  max_sec    = runtime_tab$runtime_sec[, "max"],
  total_sec  = runtime_tab$runtime_sec[, "total"]
)

print(runtime_summary, digits = 3)
```

Interpretation:

* GHK is expected to be slower than GQT.
* Misspecification may also change runtime if the likelihood becomes harder to optimize.

---

# Error messages for failed fits

If any fits failed, inspect the error messages.

```{r failed_fits}
failed_df <- fit_df[!fit_df$converged, c(
  "gen_family", "fit_family", "cell_label", "replicate", "method", "error_message"
)]

if (nrow(failed_df) == 0L) {
  cat("No failed fits.\n")
} else {
  print(failed_df)
}
```

---

# Extract parameter estimates into long format

The fitted estimates are stored in the list-column `fit_df$estimates`. We unpack them into one row per:

* generating family,
* fitting family,
* replicate,
* method,
* parameter.

We also attach a `truth` column where the comparison is well-defined.

```{r extract_estimates}
est_long <- extract_estimates_long(fit_df, fit_objects_named = fit_objects)

dim(est_long)
head(est_long)
```

---

# What does the truth mean here?

A key coding and interpretation point:

* For **Poisson-generated** data, the generation truth stores `lambda = 5`, but the fitted Poisson intercept is on the **log scale**, so the correct fitting-scale truth is:
  $$
  \log(5) \approx 1.609.
  $$

* For **NegBin-generated** data, the current config uses:
  $$
  \beta = (\log(5), 0, 0),
  $$
  so the correct intercept-scale truth is again:
  $$
  \log(5) \approx 1.609.
  $$

* The range truth is always (0.3).

* The overdispersion truth is only well-defined for the **NegBin-generated, NegBin-fitted** cell. For misspecified fits, some parameters do not have a meaningful “true” comparison, so their truth is set to `NA`.

---

# Which parameter names are present?

Before summarizing estimates, it helps to see what parameter names `gcKrig` actually returned.

```{r parameter_names}
sort(unique(est_long$parameter))
```

---

# Estimate summary by cell, method, and parameter

```{r estimate_summary}
estimate_tab <- aggregate(
  estimate ~ gen_family + fit_family + cell_label + method + parameter,
  data = est_long,
  FUN = function(x) c(
    mean   = mean(x),
    sd     = sd(x),
    median = stats::median(x),
    n      = length(x)
  )
)

estimate_summary <- data.frame(
  gen_family = estimate_tab$gen_family,
  fit_family = estimate_tab$fit_family,
  cell_label = estimate_tab$cell_label,
  method     = estimate_tab$method,
  parameter  = estimate_tab$parameter,
  mean_est   = estimate_tab$estimate[, "mean"],
  sd_est     = estimate_tab$estimate[, "sd"],
  median_est = estimate_tab$estimate[, "median"],
  n          = estimate_tab$estimate[, "n"]
)

print(estimate_summary, digits = 3)
```

Interpretation:

* In well-specified cells, the estimated means should lie near the truth.
* In misspecified cells, the estimates may drift or become more variable.

---

# MAE summary

Mean absolute error is only computed when a meaningful truth exists.

```{r mae_summary}
mae_df <- est_long[!is.na(est_long$truth), ]

mae_tab <- aggregate(
  abs_error ~ gen_family + fit_family + cell_label + method + parameter,
  data = mae_df,
  FUN = function(x) c(
    mean   = mean(x),
    median = stats::median(x),
    sd     = sd(x),
    n      = length(x)
  )
)

mae_summary <- data.frame(
  gen_family = mae_tab$gen_family,
  fit_family = mae_tab$fit_family,
  cell_label = mae_tab$cell_label,
  method     = mae_tab$method,
  parameter  = mae_tab$parameter,
  mae_mean   = mae_tab$abs_error[, "mean"],
  mae_median = mae_tab$abs_error[, "median"],
  mae_sd     = mae_tab$abs_error[, "sd"],
  n          = mae_tab$abs_error[, "n"]
)

print(mae_summary, digits = 3)
```

Interpretation:

* In the well-specified cells, MAE is the main accuracy metric.
* In misspecified cells, MAE is still interpretable for shared parameters like range and intercept-scale mean, but not always for parameters that only exist under one family.

---

# Information criteria summary

AIC/BIC/AICc are recorded by the wrapper. These can be useful descriptively, but they should be interpreted carefully across misspecified families.

```{r ic_summary}
ic_df <- fit_df[fit_df$converged, ]

ic_tab <- aggregate(
  cbind(loglik, aic, bic, aicc) ~ gen_family + fit_family + cell_label + method,
  data = ic_df,
  FUN = mean
)

print(ic_tab, digits = 3)
```

---

# Boxplots of runtime

If `ggplot2` is available, produce a clean runtime plot. Otherwise use base graphics.

```{r runtime_plot}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  ggplot(runtime_df, aes(x = method, y = runtime_sec, fill = method)) +
    geom_boxplot(outlier.alpha = 0.5) +
    facet_wrap(~ cell_label, scales = "free_y") +
    labs(
      title = "Runtime by method across Day 3 cells",
      x = "Method",
      y = "Runtime (seconds)"
    ) +
    theme_minimal()
} else {
  boxplot(
    runtime_sec ~ interaction(cell_label, method),
    data = runtime_df,
    las = 2,
    main = "Runtime by cell and method",
    ylab = "Runtime (seconds)",
    xlab = ""
  )
}
```

---

# Boxplots of parameter estimates

Again, use `ggplot2` if available.

```{r estimate_plot}
if (nrow(est_long) > 0) {
  if (requireNamespace("ggplot2", quietly = TRUE)) {
    library(ggplot2)

    ggplot(est_long, aes(x = method, y = estimate, fill = method)) +
      geom_boxplot(outlier.alpha = 0.5) +
      facet_grid(parameter ~ cell_label, scales = "free_y") +
      geom_hline(aes(yintercept = truth), colour = "red", linetype = 2, na.rm = TRUE) +
      labs(
        title = "Estimate distributions across Day 3 misspecification cells",
        x = "Method",
        y = "Estimate"
      ) +
      theme_minimal()
  } else {
    par(mfrow = c(1, 1))
    boxplot(
      estimate ~ interaction(method, cell_label),
      data = est_long,
      las = 2,
      main = "Estimate distributions",
      ylab = "Estimate",
      xlab = ""
    )
  }
}
```

Interpretation:

* The dashed red line is the truth where defined.
* In well-specified cells, boxes should be centered closer to the red line.
* In misspecified cells, the distributions may shift or widen.

---

# A simple 2 x 2 Day 3 summary table

This is the most report-like table in the document.

It focuses on:

* convergence,
* mean runtime,
* mean MAE for the parameters where truth exists.

```{r day3_summary_table}
# convergence
conv_small <- convergence_summary[, c("cell_label", "method", "conv_rate")]

# runtime
rt_small <- runtime_summary[, c("cell_label", "method", "mean_sec")]

# MAE
mae_small <- mae_summary[, c("cell_label", "method", "parameter", "mae_mean")]

# merge convergence + runtime first
cell_method_tab <- merge(conv_small, rt_small, by = c("cell_label", "method"), all = TRUE)

# reshape MAE wider by parameter
if (nrow(mae_small) > 0) {
  mae_wide <- reshape(
    mae_small,
    idvar = c("cell_label", "method"),
    timevar = "parameter",
    direction = "wide"
  )

  day3_summary <- merge(cell_method_tab, mae_wide, by = c("cell_label", "method"), all = TRUE)
} else {
  day3_summary <- cell_method_tab
}

print(day3_summary, digits = 3)
```

---

# Interpretation guide

At this stage, the most important comparisons are:

## 1. Well-specified versus misspecified

Compare:

* Poisson ($\rightarrow$) Poisson versus Poisson ($\rightarrow$) NegBin
* NegBin ($\rightarrow$) NegBin versus NegBin ($\rightarrow$) Poisson

If the misspecified cell has:

* worse convergence,
* slower runtime,
* larger MAE,
* more variable estimates,

then the misspecification study is already producing a meaningful result.

## 2. GHK versus GQT under misspecification

The Day 1 and Day 2 baseline focused on GHK versus GQT under correct specification.

Day 3 asks a more robust question:

> when the fitted marginal family is wrong, does GQT still behave similarly to GHK, or does one method degrade more severely?

The key summaries for that are:

* convergence rates,
* runtime,
* MAE,
* estimate boxplots.

## 3. Shared-parameter interpretation

Some parameters remain comparable even under misspecification:

* the intercept-scale mean parameter,
* the spatial range.

Other parameters are only meaningful under one family, especially overdispersion.

So the cleanest Day 3 interpretation usually focuses first on:

* range,
* intercept-scale mean,
* convergence,
* runtime.

---

# Final takeaway

The Day 3 misspecification pipeline is now doing exactly what it should:

1. generate controlled Poisson and NegBin spatial count datasets,
2. fit each under both correct and incorrect marginal assumptions,
3. store one fit-result object per cell,
4. summarize how misspecification affects stability, speed, and accuracy.

This document is the analysis layer on top of that pipeline.

---
