---
title: "Missingness End-to-End: Generation, Masking, Fitting, Inspection"
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  message = TRUE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
```

## Purpose

Walk through the full missingness pipeline interactively. Tests the
masking logic, the NA-aware fitting, and inspection of a single
scenario before committing to the full Slurm run.

**Safety convention.** The full 5-scenario x 2-method x 200-replicate
run is marked `eval = FALSE` — Knit completes quickly, and the
expensive chunk is only executed when you click the green arrow.
All the smoke-test chunks do run on Knit.

---

## Background: what this study does

**Goal.** Test the robustness of gcKrig estimators under Missing
Completely At Random (MCAR) missingness. Take the K13-A baseline
(Poisson(5), n=49, range=0.3) and randomly mask a fraction of
observations per replicate, then fit. Compare the resulting MAE and
bias to the unmasked baseline.

**Why MCAR specifically.** Under MCAR, the decision of which
observations are missing is independent of their values. Estimators
remain theoretically unbiased; only their variance grows. This
gives us a clean test of how gracefully gcKrig handles reduced
effective sample size. (Non-random missingness — e.g., masking
large values preferentially — would be a different study that
deliberately violates the MCAR assumption.)

**Scenarios.** Five missingness rates: 0%, 10%, 20%, 30%, 50%. All
share the same underlying simulated data (same seed); only the
mask differs. This keeps the comparison honest: observed
differences in MAE arise purely from data availability, not from
sampling variability in the generating process.

**Expected findings.**

| Quantity | Expectation |
|----------|-------------|
| Convergence rate | ~100% at all rates up to 50% |
| Bias | Near zero at all rates (MCAR property) |
| MAE, rate=0 | Matches K13-A default baseline |
| MAE growth | Roughly linear in rate, moderate even at 50% |
| Runtime | Decreases with rate (smaller covariance matrix) |
| GHK vs GQT | Degrade at similar rates; no clear winner |

If results deviate sharply from these expectations (e.g., bias
grows with rate, or convergence breaks down above 30%), that itself
becomes a finding worth discussing in the report.

---

## Load the pipeline
test
```{r test}
source("code/R/logging.R"); source("code/R/io.R"); source("code/R/fit_core.R")
source("code/R/grids.R"); source("code/R/spec_utils.R"); source("code/R/sim_core.R")
source("code/R/missingness.R"); source("code/configs/missing.R")
library(gcKrig)

# Generate a quick test scenario
spec <- MISSING_SCENARIOS$mcar_rate50
locs <- do.call(make_locations, spec$layout)
base_scen <- simulate_mc_scenario(spec = spec, locations = locs,
  marginal = poisson.gc(lambda = 5),
  corr = matern.gc(range = 0.3, kappa = 0.5, nugget = 0))
masked <- apply_mcar(base_scen, missing_rate = 0.5)

# Fit 2 replicates
fit_scenario(masked, replicates = 1:2)
```

```{r load_pipeline}
suppressPackageStartupMessages({
  library(gcKrig)
})

source("code/R/logging.R")
source("code/R/grids.R")
source("code/R/spec_utils.R")
source("code/R/sim_core.R")
source("code/R/io.R")
source("code/R/missingness.R")
source("code/R/fit_core.R")
source("code/configs/missing.R")
```

### Peek at the scenarios list

```{r show_scenarios}
names(MISSING_SCENARIOS)

# Inspect one:
str(MISSING_SCENARIOS$mcar_rate30, max.level = 2)
```

Note that the rate is stored in `$missing_rate`, and all five
scenarios share the same `$seed` (20240601). The rate governs how
much of the simulated data gets masked; the seed governs the
underlying simulation, which is identical across all five.

---

## Part 1: Generate a test scenario

### Simulate the baseline first (no masking yet)

```{r generate_baseline}
spec <- MISSING_SCENARIOS$mcar_rate30   # 30% missingness

locs <- do.call(make_locations, spec$layout)

marginal_obj <- poisson.gc(lambda = spec$marginal$lambda)
corr_obj     <- matern.gc(range  = spec$corr$range,
                          kappa  = spec$corr$kappa,
                          nugget = spec$corr$nugget)

scenario_unmasked <- simulate_mc_scenario(
  spec        = spec,
  locations   = locs,
  marginal    = marginal_obj,
  corr        = corr_obj,
  true_params = list(lambda = 5, range = 0.3, kappa = 0.5, nugget = 0)
)

cat("Unmasked shape:", dim(scenario_unmasked$values), "\n")
cat("NA count:      ", sum(is.na(scenario_unmasked$values)), "\n")
cat("Sample values (replicate 1):\n")
print(scenario_unmasked$values[1:10, 1])
```

### Apply MCAR masking

```{r apply_masking}
scenario_masked <- apply_mcar(scenario_unmasked, missing_rate = 0.3)

cat("Masked shape:   ", dim(scenario_masked$values), "\n")
cat("NA count:       ", sum(is.na(scenario_masked$values)), "\n")
cat("Fraction NA:    ", round(mean(is.na(scenario_masked$values)), 3), "\n")
cat("Full preserved: ", !is.null(scenario_masked$values_full), "\n")
cat("\nSample values (replicate 1), masked:\n")
print(scenario_masked$values[1:10, 1])
cat("\nSame replicate, original:\n")
print(scenario_masked$values_full[1:10, 1])
```

Expected: the two vectors above differ exactly in positions where
the mask set a value to NA. Everything else is identical. The
scenario now carries both `$values` (with NAs) and `$values_full`
(the pre-masking original), plus a `$missingness_spec` record:

```{r missingness_spec}
str(scenario_masked$missingness_spec)
```

---

## Part 2: Visualize the masking pattern

### Compare spatial patterns: all 49 vs. 30% masked

```{r plot_masking, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))

v_full <- scenario_masked$values_full[, 1]
v_mask <- scenario_masked$values[, 1]

plot(
  scenario_masked$locations[, "x"],
  scenario_masked$locations[, "y"],
  pch  = 19,
  cex  = 0.5 + 2 * v_full / max(v_full),
  col  = hcl.colors(50, "YlOrRd", rev = TRUE)[
    cut(v_full, breaks = 50, labels = FALSE)
  ],
  xlab = "x", ylab = "y",
  main = "Complete data (n=49)"
)

# Grey out missing points
col_mask <- ifelse(is.na(v_mask), "grey70",
  hcl.colors(50, "YlOrRd", rev = TRUE)[cut(v_full, breaks = 50, labels = FALSE)])
cex_mask <- ifelse(is.na(v_mask), 1.5, 0.5 + 2 * v_full / max(v_full))
pch_mask <- ifelse(is.na(v_mask), 4, 19)

plot(
  scenario_masked$locations[, "x"],
  scenario_masked$locations[, "y"],
  pch  = pch_mask,
  cex  = cex_mask,
  col  = col_mask,
  xlab = "x", ylab = "y",
  main = "Observed data (n=34, 30% masked)"
)

par(mfrow = c(1, 1))
```

The left panel shows the full generating process; the right panel
shows what a fit would actually see. X marks are the masked points.

---

## Part 3: Smoke-test fitting

Before launching the full study, fit the masked scenario on 1
replicate with each method. Verifies that `.fit_one()` correctly
drops NA rows and that `n_used` reports the reduced sample size.

```{r smoke_fit}
log_info("Smoke test: fitting masked scenario, replicate 1, both methods.")

smoke <- fit_scenario(
  scenario   = scenario_masked,
  methods    = c("GHK", "GQT"),
  replicates = 1
)
```

```{r smoke_inspect}
cat("Results for replicate 1:\n")
print(smoke$results[, c("method", "converged", "runtime_sec",
                        "n_used", "n_full", "error_message")])

cat("\nGHK estimates:\n")
print(smoke$results$estimates[[1]])
cat("\nGQT estimates:\n")
print(smoke$results$estimates[[2]])
cat("\nTrue values: log(5) =", log(5), ", range = 0.3\n")
```

**What to check:**

- `converged` should be TRUE for both rows.
- `n_used` should be around 34 (49 × 0.7), reflecting that 30% were
  masked. If it's 49, the masking isn't propagating to the fitter.
- Estimates should be near the truth. Single-replicate noise is
  normal; the Intercept should be in the 1.4 to 1.8 range,
  range should be in the 0.15 to 0.45 range.

---

## Part 4: Compare masked vs unmasked fits on the same replicate

Worth seeing directly: does losing 30% of the data change the
estimates much?

```{r compare_masked_unmasked}
log_info("Fitting the SAME replicate with and without masking.")

smoke_full <- fit_scenario(
  scenario   = scenario_unmasked,
  methods    = c("GHK", "GQT"),
  replicates = 1
)

# Compare estimates side by side
compare_tab <- data.frame(
  method   = rep(c("GHK", "GQT"), 2),
  scenario = rep(c("complete", "masked30"), each = 2),
  intercept = c(
    smoke_full$results$estimates[[1]]$Intercept,
    smoke_full$results$estimates[[2]]$Intercept,
    smoke$results$estimates[[1]]$Intercept,
    smoke$results$estimates[[2]]$Intercept
  ),
  range = c(
    smoke_full$results$estimates[[1]]$range,
    smoke_full$results$estimates[[2]]$range,
    smoke$results$estimates[[1]]$range,
    smoke$results$estimates[[2]]$range
  ),
  n_used = c(
    smoke_full$results$n_used[1:2],
    smoke$results$n_used[1:2]
  ),
  runtime_sec = round(c(
    smoke_full$results$runtime_sec[1:2],
    smoke$results$runtime_sec[1:2]
  ), 3)
)

print(compare_tab, row.names = FALSE, digits = 4)
cat("\nTruth: Intercept =", round(log(5), 4), "  range = 0.3\n")
```

Single-replicate comparison isn't meant to be statistically
conclusive — the real comparison comes from the full study with 200
replicates per rate. But if the masked estimates are *wildly*
different from the unmasked ones (e.g., sign-flipped, off by an
order of magnitude), something is wrong and we should debug before
proceeding.

---

## Part 5: Full pipeline run (heavy — `eval = FALSE` by default)

This is the full 5-scenario × 200-replicate × 2-method run: 2,000
fits total. Expected runtime 15-35 minutes interactively, or launch
via Slurm for the same work without tying up the R session.

### Option A: run it here in the Rmd

```{r full_run_local, eval = FALSE}
log_info("Full missingness generation: 5 scenarios")

for (scen_id in names(MISSING_SCENARIOS)) {
  spec <- MISSING_SCENARIOS[[scen_id]]

  locs <- do.call(make_locations, spec$layout)
  scn  <- simulate_mc_scenario(
    spec      = spec,
    locations = locs,
    marginal  = poisson.gc(lambda = spec$marginal$lambda),
    corr      = matern.gc(range = spec$corr$range,
                          kappa = spec$corr$kappa,
                          nugget = spec$corr$nugget),
    true_params = list(lambda = 5, range = 0.3, kappa = 0.5, nugget = 0)
  )

  if (spec$missing_rate > 0) {
    scn <- apply_mcar(scn, missing_rate = spec$missing_rate)
  }

  save_scenario(scn)
}

log_info("Full missingness fitting: 5 scenarios")

for (scen_id in names(MISSING_SCENARIOS)) {
  scn <- load_scenario("MISSING", scen_id)
  fit <- fit_scenario(scn, methods = c("GHK", "GQT"))
  save_fit_result(fit)
}
```

### Option B: run it via Slurm (recommended)

```bash
sbatch code/slurm/05_gen_missing.slurm
# wait for it to finish
sbatch code/slurm/13_fit_missing.slurm
```

Either way produces the same files. Option B is better when you
don't want to occupy an R session for 30+ minutes.

---

## Part 6: Inspect results

Once the full run has completed (whichever path), load results and
check them. See `missing_analysis.Rmd` for the full analysis; this
section is a quick sanity check.

```{r inspect_results}
avail <- list_fit_results("MISSING")

if (nrow(avail) > 0) {
  cat("Available fit results:\n")
  print(avail, row.names = FALSE)

  # Load rate=0 and rate=50 for quick contrast
  fr0  <- load_fit_result("MISSING", "mcar_rate0")
  fr50 <- load_fit_result("MISSING", "mcar_rate50")

  cat("\n=== rate=0 (no masking) ===\n")
  cat("Converged:", sum(fr0$results$converged), "/", nrow(fr0$results), "\n")
  cat("Mean n_used:", mean(fr0$results$n_used), "\n")

  cat("\n=== rate=50 ===\n")
  cat("Converged:", sum(fr50$results$converged), "/", nrow(fr50$results), "\n")
  cat("Mean n_used:", mean(fr50$results$n_used), "\n")

} else {
  message("No MISSING fit results yet. Run the full pipeline first.")
}
```

**Sanity signals.**

- rate=0 should show mean n_used = 49 exactly.
- rate=50 should show mean n_used around 24-25.
- Convergence should be near 100% for both.
- If rate=50 shows materially degraded convergence, flag that as a
  finding and investigate whether it's systemic or a handful of
  pathological replicates.

---

## Next steps

Once this runs cleanly end to end, the next extension is
**zero-inflation** — generating data from a ZIP marginal and fitting
as both ZIP (well-specified) and Poisson (misspecified). The
infrastructure for that already exists via the misspecification
plumbing in `fit_core.R`; the new code is a config file and a
generation script, much like this missingness study.

After that, **irregular sampling** — replacing the regular grid
layout with random or clustered location patterns. Requires
extending `make_locations()` with new branches but nothing
fundamentally new at the fitter level.

For analysis, see `missing_analysis.Rmd`, which reads the five fit
result files and produces the tables and figures for the report.
