---
title: "STAT778 Project Codebase Guide: Simulation, Fitting, and Results"
author: "STAT778 Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Purpose of this document

This document is a code-level guide to the current STAT778 project pipeline. It is written from the **coding standpoint first**, and it inserts theory only where the theory helps explain what the code is doing.

The main goals are:

1. Explain how simulation scenarios are defined.
2. Explain how simulated data are generated.
3. Explain how fitted models are run.
4. Explain how outputs are stored and interpreted.
5. Clarify what parameters and settings go into the code.
6. Clarify the difference between **generation-scale truth** and **fitting-scale truth**.

This document is meant to answer questions like:

* What exactly is a scenario in this codebase?
* What is the exact input to the simulation wrapper?
* What is the exact input to the fitting wrapper?
* What does the fitting wrapper return?
* Why do fitted parameters sometimes appear on a different scale than the simulation truth?
* What are the four scenario families and what role does each play?

---

# Repository structure

At a high level, the repository is organized around:

* a shared library under `code/R/`
* per-dataset configs under `code/configs/`
* per-dataset driver scripts under `code/scripts/`
* Slurm submission scripts under `code/slurm/`
* saved outputs under `data/generated/` and `data/fits/`

A typical layout is:

```text
STAT778_Project/
├── README.md
├── h18b_example.Rmd
├── code/
│   ├── R/
│   │   ├── logging.R
│   │   ├── grids.R
│   │   ├── spec_utils.R
│   │   ├── sim_core.R
│   │   ├── fit_core.R
│   │   ├── io.R
│   │   └── helper_functions.R
│   ├── configs/
│   │   ├── h18b.R
│   │   ├── k13a.R
│   │   ├── k13b.R
│   │   └── k13c.R
│   ├── scripts/
│   │   ├── 01_gen_h18b.R
│   │   ├── 02_gen_k13a.R
│   │   ├── ...
│   │   └── 10_fit_k13a.R
│   └── slurm/
│       ├── _env.sh
│       ├── 01_gen_h18b.slurm
│       ├── 02_gen_k13a.slurm
│       └── 10_fit_k13a.slurm
├── data/
│   ├── generated/
│   └── fits/
├── logs/
└── visualization/
```

The design principle is simple:

* **configs** define what should exist,
* **scripts** execute one pipeline stage,
* **shared R files** contain reusable logic,
* **saved objects** are the persistent interface between stages.

---

# The four scenario families

The project currently works with four scenario families.

## H18-B

H18-B is the Han and De Oliveira (2018)-style warm-up scenario. It is mainly used to verify that the simulation and fitting environment works correctly inside `gcKrig`.

Its role in the project is:

* sanity check for installation and package setup,
* first end-to-end pipeline test,
* package-comparison warm-up.

## K13-A

K13-A is the main Kazianka (2013) estimation study scenario family. It is a Monte Carlo simulation design with repeated replicated datasets under one parameter cell.

Its role in the project is:

* the main baseline estimation study,
* comparison of GHK and GQT,
* the launch point for later misspecification and robustness extensions.

## K13-B

K13-B is the Poisson prediction-accuracy benchmark. It is not a large Monte Carlo design. Instead, it is a fixed benchmark geometry with fixed observed values and a parameter grid over which predictive distributions are compared.

Its role in the project is:

* isolating approximation error in predictive distributions,
* comparing exact or benchmark prediction against approximate prediction.

## K13-C

K13-C is the Bernoulli analogue of K13-B. It uses fixed binary observation patterns on the same simple cross-shaped geometry.

Its role in the project is:

* stress-testing prediction in the highly discrete regime,
* understanding where approximation becomes unstable.

So there are “four scenario families,” but they are not all the same kind of object:

* **H18-B and K13-A** are Monte Carlo simulation scenarios.
* **K13-B and K13-C** are fixed benchmark scenarios for prediction.

---

# Shared design idea: the scenario object

A central idea in this codebase is that each dataset is represented by a **scenario object**.

A scenario object is a standard list that stores:

* identifiers (`dataset_id`, `scenario_id`)
* the full generating config
* spatial locations
* simulated values or fixed observed values
* true generating parameters
* provenance fields such as seed, package version, and creation time

This means the project always tries to save **self-describing objects**. Once a scenario has been saved to disk, it should be possible to load it and know:

* where it came from,
* what parameters generated it,
* how many replicates it contains,
* what the fitting code should do with it.

---

# Shared design idea: the fit-result object

Fitting does not overwrite scenario objects. Instead, fitting produces a separate **fit-result object**.

A fit-result object stores:

* identifiers (`dataset_id`, `scenario_id`)
* the scenario config that was fit
* a long-format results table
* provenance fields such as time and package version

This keeps generation and fitting cleanly separated:

* scenario objects represent **inputs to fitting**
* fit-result objects represent **outputs from fitting**

---

# The full code pipeline in one sentence

The full pipeline is:

1. a config file defines one or more scenarios,
2. a script generates and saves scenario objects,
3. another script fits those scenario objects and saves fit-result objects,
4. an RMarkdown analysis file loads those saved objects and produces summaries, plots, and interpretations.

---

# Load the shared pipeline

```{r load_pipeline}
suppressPackageStartupMessages({
  library(gcKrig)
})

source("code/R/logging.R")
source("code/R/grids.R")
source("code/R/spec_utils.R")
source("code/R/sim_core.R")
source("code/R/fit_core.R")
source("code/R/io.R")
```

Optionally load any configs that exist locally:

```{r load_configs}
if (file.exists("code/configs/h18b.R")) source("code/configs/h18b.R")
if (file.exists("code/configs/k13a.R")) source("code/configs/k13a.R")
if (file.exists("code/configs/k13b.R")) source("code/configs/k13b.R")
if (file.exists("code/configs/k13c.R")) source("code/configs/k13c.R")
```

---

# Part I: Simulation pipeline

# 1. What is the exact input to the simulation wrapper?

For Monte Carlo simulation scenarios like H18-B and K13-A, the simulation wrapper receives five main inputs:

1. `spec`
   A nested list containing identifiers, layout, marginal spec, correlation spec, seed, replicate count, and metadata.

2. `locations`
   A numeric matrix of coordinates, usually `n x 2`.

3. `marginal`
   A `marginal.gc` object from `gcKrig`, such as:

   * `poisson.gc(lambda = 5)`
   * `negbin.gc(mu = ..., od = 1)`

4. `corr`
   A `corr.gc` object from `gcKrig`, such as:

   * `matern.gc(range = 0.3, kappa = 0.5, nugget = 0)`

5. `true_params`
   A flat named list used only for bookkeeping and later comparison.

Then the simulation wrapper calls `gcKrig::simgc()` and standardizes the output into the project schema.

---

# 2. Where do the scenario specs come from?

They come from config files under `code/configs/`.

For example, the current K13-A pilot should be defined in `code/configs/k13a.R`. If that config exists, you can inspect it directly.

```{r inspect_k13a_spec, eval = exists("K13A_SCENARIOS")}
if (exists("K13A_SCENARIOS")) {
  names(K13A_SCENARIOS)
  str(K13A_SCENARIOS$default, max.level = 3)
}
```

The important point is:

> The config is a declarative recipe. It says what the scenario should be, but it does not itself generate any data.

---

# 3. How locations are generated

The file `code/R/grids.R` is responsible for constructing the spatial layout.

For K13-A, the default pilot uses a regular grid. The config says “regular grid with side length 7,” and the code turns that into a `49 x 2` location matrix.

```{r build_locations, eval = exists("K13A_SCENARIOS")}
if (exists("K13A_SCENARIOS")) {
  spec <- K13A_SCENARIOS$default
  locs <- do.call(make_locations, spec$layout)

  dim(locs)
  head(locs)
  tail(locs)
}
```

For a regular grid, the row order matters because later:

* rows of `locations`,
* rows of `values`

must correspond exactly.

---

# 4. Theory behind the simulation

The simulations are based on **Gaussian copula spatial models**.

At the theory level, the construction is:

1. Generate a latent Gaussian random field
   $$
   Z = (Z(s_1), \dots, Z(s_n))
   $$
   over the chosen locations, with mean zero, unit variance, and a spatial correlation matrix determined by the chosen correlation model.

2. Convert the latent Gaussian values to uniforms:
   $$
   U_i = \Phi(Z(s_i)).
   $$

3. Convert uniforms to the target marginal distribution:
   $$
   Y(s_i) = F^{-1}(U_i).
   $$

This gives discrete spatial data whose:

* marginal distribution is controlled by $F$,
* dependence structure is controlled by the latent Gaussian field.

This is the core idea used in both Kazianka (2013) and gcKrig.

---

# 5. K13-A default: exact simulation setup

For the current default pilot, the setup is:

* $n = 49$ locations on a $7 \times 7$ regular grid,
* `sim_n = 30` independent Monte Carlo replicates,
* marginal family: Poisson,
* marginal mean: $\lambda = 5$,
* correlation family: Matérn,
* range: $0.3$,
* smoothness: $\kappa = 0.5$,
* nugget: $0$.

Because $\kappa = 0.5$, the Matérn correlation reduces to the exponential model:
$$
\rho(h) = \exp(-h / r),
$$
with $r = 0.3$.

So each replicate is a length-49 vector of correlated Poisson counts.

---

# 6. What `simulate_mc_scenario()` returns

The simulation wrapper returns a **scenario object** created by `make_scenario()`.

Let us load the generated default K13-A object if it exists.

```{r load_generated_scenario, eval = file.exists("data/generated/K13-A/default.rds")}
if (file.exists("data/generated/K13-A/default.rds")) {
  scenario <- load_scenario("K13-A", "default")
  str(scenario, max.level = 1)
}
```

The most important fields are:

* `scenario$locations`: the `n x 2` coordinate matrix,
* `scenario$values`: the `n x B` count matrix,
* `scenario$true_params`: generation truth,
* `scenario$config`: the original spec.

```{r inspect_generated_values, eval = exists("scenario")}
if (exists("scenario")) {
  dim(scenario$locations)
  dim(scenario$values)
  scenario$true_params
  scenario$seed_info
}
```

Interpretation of `values`:

* each **row** corresponds to one spatial location,
* each **column** corresponds to one Monte Carlo replicate.

So:

```{r replicate_vector, eval = exists("scenario")}
if (exists("scenario")) {
  y1 <- scenario$values[, 1]
  length(y1)
  head(y1)
}
```

is the full first dataset to be fit.

---

# 7. Quick visualization of one simulated replicate

```{r plot_one_replicate, eval = exists("scenario")}
if (exists("scenario")) {
  rep_i <- 1
  values_i <- scenario$values[, rep_i]

  plot(
    scenario$locations[, "x"],
    scenario$locations[, "y"],
    pch = 19,
    cex = 0.5 + 2 * values_i / max(values_i),
    col = hcl.colors(50, "YlOrRd", rev = TRUE)[
      cut(values_i, breaks = 50, labels = FALSE)
    ],
    xlab = "x", ylab = "y",
    main = sprintf("K13-A default, replicate %d", rep_i)
  )
}
```

This is often the fastest way to check that:

* the scenario loaded correctly,
* the values are not degenerate,
* there is visible spatial structure.

---

# 8. What do the “true parameters” mean?

The `true_params` field stores the **generation-scale truth**, not necessarily the fitting-scale truth.

For the current K13-A pilot, it stores things like:

* `lambda = 5`
* `range = 0.3`
* `kappa = 0.5`
* `nugget = 0`

These are the parameters that generated the data.

```{r show_true_params, eval = exists("scenario")}
if (exists("scenario")) {
  scenario$true_params
}
```

But the fitting code does **not** necessarily estimate all of these directly.

That is a major source of confusion, so the next section explains it carefully.

---

# Part II: Fitting pipeline

# 9. The key idea of fitting

The fitting wrapper does **not** fit the scenario object as one big dataset.

Instead, it works **one replicate at a time**.

If the scenario contains a `49 x 30` matrix, then there are 30 separate datasets inside it, and each one must be fit separately.

The fitting wrapper automates that process.

---

# 10. The three main layers in `fit_core.R`

The file `code/R/fit_core.R` has three important levels.

## Layer 1: `.fit_one()`

This fits **one replicate** with **one method**.

Inputs:

* one scenario object,
* one replicate index,
* one fitting method (`"GHK"` or `"GQT"`).

Output:

* one one-row data frame containing fit metadata and results.

## Layer 2: `fit_scenario()`

This loops over all requested replicates and methods.

For example, if there are 30 replicates and 2 methods, it produces:
$$
30 \times 2 = 60
$$
rows of results.

## Layer 3: `make_fit_result()`

This packages the final output into a standard fit-result object.

So the fitting pipeline is:

* **input:** one scenario object,
* **work:** fit every requested replicate by every requested method,
* **output:** one fit-result object with a long-format results table.

---

# 11. What is the exact input to `.fit_one()`?

For one replicate, `.fit_one()` extracts:

```r
y    <- scenario$values[, replicate]
locs <- scenario$locations
spec <- scenario$config
```

So the real data passed to `mlegc()` are:

* one count vector `y`,
* one location matrix `locs`.

Then it rebuilds the gcKrig objects **for fitting**, not for simulation.

This distinction matters.

---

# 12. Why fitting objects differ from simulation objects

For simulation, the marginal and correlation objects were built with the true parameters filled in.

For example:

* `poisson.gc(lambda = 5)`
* `matern.gc(range = 0.3, kappa = 0.5, nugget = 0)`

For fitting, the model must be told **what family to assume**, but the parameters to be estimated should be **left unset**.

So the fitting builder does things like:

* `poisson.gc()`
* `negbin.gc()`
* `binomial.gc(size = 1)`
* `matern.gc(kappa = ..., nugget = ...)`

This tells `mlegc()` what type of model to fit.

---

# 13. What gets estimated in the default K13-A fit?

This is a critical point.

The generation truth stores:

* `lambda = 5`
* `range = 0.3`
* `kappa = 0.5`
* `nugget = 0`

But the fitting code does **not** estimate all four.

For the default pilot:

* the Poisson marginal parameter is estimated,
* the range is estimated,
* `kappa` is held fixed at `0.5`,
* nugget is fixed at zero and not estimated.

So the fit is estimating a **smaller parameter set** than the full simulation truth.

This is by design.

---

# 14. Why the fitted Poisson parameter appears as 1.6 instead of 5

gcKrig fits the Poisson mean using a **log link**.

That means the fitted marginal parameter is not $\lambda$ directly. It is the intercept:
$$
\beta_0 = \log(\lambda).
$$

For the default truth $\lambda = 5$:
$$
\beta_0^{\text{true}} = \log(5) \approx 1.609.
$$

So:

* generation-scale truth: `lambda = 5`,
* fitting-scale truth: `Intercept = log(5) ≈ 1.609`.

That is why boxplots may show a truth line near `1.6`.

Both refer to the same underlying truth, just on different scales.

---

# 15. Smoke test: fit one replicate with both methods

The easiest way to understand the wrapper is to use it on one replicate.

```{r smoke_fit, eval = exists("scenario")}
if (exists("scenario")) {
  smoke_result <- fit_scenario(
    scenario   = scenario,
    methods    = c("GHK", "GQT"),
    replicates = 1
  )

  str(smoke_result, max.level = 1)
  smoke_result$results
}
```

This asks the wrapper to do only:

* replicate 1 with GHK,
* replicate 1 with GQT.

So the results table should have exactly 2 rows.

---

# 16. What the fit-result object contains

The output of `fit_scenario()` is a fit-result object with:

* `dataset_id`
* `scenario_id`
* `scenario_config`
* `results`
* provenance fields like creation time and package version

The key component is `fit_result$results`.

```{r inspect_fit_result, eval = exists("smoke_result")}
if (exists("smoke_result")) {
  res <- smoke_result$results
  names(res)
  res
}
```

Typical columns are:

* `replicate`
* `method`
* `converged`
* `error_message`
* `runtime_sec`
* `loglik`
* `aic`
* `bic`
* `aicc`
* `estimates`
* `std_errors`

The last two are list-columns. Each row has its own estimate list and standard-error list.

---

# 17. Looking inside one fitted row

```{r inspect_one_fit_row, eval = exists("smoke_result")}
if (exists("smoke_result")) {
  res <- smoke_result$results
  res$estimates[[1]]
  res$std_errors[[1]]
}
```

This is usually where you see parameter names such as:

* `Intercept`
* `range`

Those are the fitted parameter names returned by gcKrig.

For the current pilot, the correct truth comparisons are:

* `Intercept` versus `log(5)`,
* `range` versus `0.3`.

---

# 18. What “converged” means in this codebase

In the wrapper, `converged = TRUE` means:

* `mlegc()` returned without throwing an error.

`converged = FALSE` means:

* the fit failed,
* the wrapper caught the error,
* the error message was stored instead of stopping the whole scenario.

This is one of the main reasons to have the wrapper at all. In a large simulation study, one bad replicate should not crash the whole run.

---

# Part III: Results and evaluation

# 19. What is evaluated from the fit results?

The project currently evaluates methods through summaries of the `results` table.

The main quantities are:

* convergence rate by method,
* runtime by method,
* distributions of fitted parameters,
* mean absolute error (MAE) versus truth.

---

# 20. Convergence summary

```{r convergence_summary, eval = exists("smoke_result")}
if (exists("smoke_result")) {
  res <- smoke_result$results
  tab <- table(method = res$method, converged = res$converged)
  tab
  prop.table(tab, margin = 1)
}
```

For the smoke test this is not very informative, but for the full 30-replicate fit it becomes meaningful.

---

# 21. Runtime summary

```{r runtime_summary, eval = exists("smoke_result")}
if (exists("smoke_result")) {
  res <- smoke_result$results
  aggregate(runtime_sec ~ method, data = res, mean)
}
```

For the full fit-result object this tells you how much slower GHK is than GQT.

---

# 22. Estimate extraction and truth comparison

For full evaluation, the `estimates` list-column is unpacked into long format.

The logic is:

1. for each row of the results table,
2. take the named estimate list,
3. make one row per parameter,
4. attach the truth value,
5. compute error summaries such as MAE.

The crucial subtlety is that truth must be compared on the **correct scale**:

* `Intercept` should be compared to `log(lambda)`,
* `range` should be compared to the stored true range.

---

# 23. Full fit object on disk

Once a full fit is run and saved, you can reload it with:

```{r load_saved_fit, eval = FALSE}
fit_result <- load_fit_result("K13-A", "default")
str(fit_result, max.level = 1)
dim(fit_result$results)
```

If all 30 replicates are fit by both methods, then:

* `fit_result$results` should have 60 rows.

---

# Part IV: How to think about the whole pipeline

# 24. End-to-end summary

The codebase pipeline is:

## Step 1: config

A config file defines a scenario spec.

## Step 2: generate

A generation script:

* builds locations,
* builds gcKrig simulation objects,
* calls `simulate_mc_scenario()`,
* saves a scenario object.

## Step 3: fit

A fitting script:

* loads a scenario object,
* calls `fit_scenario()`,
* saves a fit-result object.

## Step 4: analyze

An RMarkdown or analysis script:

* loads scenario and fit-result objects,
* summarizes convergence, timing, estimates, and errors,
* produces tables and plots.

So the project is really built around two persistent object types:

* **scenario objects**
* **fit-result objects**

---

# 25. Practical inspection checklist

When something feels confusing, these are the most useful object-level checks.

## For a scenario object

```{r scenario_checklist, eval = exists("scenario")}
if (exists("scenario")) {
  dim(scenario$locations)
  dim(scenario$values)
  scenario$true_params
  head(scenario$values[, 1])
}
```

## For a fit-result object

```{r fit_checklist, eval = exists("smoke_result")}
if (exists("smoke_result")) {
  smoke_result$dataset_id
  smoke_result$scenario_id
  dim(smoke_result$results)
  smoke_result$results[, c("replicate", "method", "converged", "runtime_sec")]
  smoke_result$results$estimates[[1]]
}
```

These usually tell you:

* what went in,
* what came out,
* what scale the parameters are on.

---

# 26. Recommended workflow

The intended workflow is:

1. **Use RMarkdown interactively** for:

   * small-scale inspection,
   * smoke-test fitting,
   * checking object structures,
   * developing plots and summaries.

2. **Use Slurm** for:

   * large replicate counts,
   * many scenario cells,
   * full fitting runs.

3. **Return to RMarkdown** to:

   * read saved results,
   * compute MAEs,
   * make report figures,
   * write interpretation.

This separation keeps debugging fast and production runs reproducible.

---

# 27. Final takeaway

From a coding standpoint, the project has a very simple architecture once the wrappers are understood:

* the **simulation wrapper** turns a config plus gcKrig objects into a saved scenario object,
* the **fitting wrapper** turns a scenario object into a saved fit-result object,
* the **analysis layer** reads those saved objects and computes summaries.

The theory only enters at a few places:

* latent Gaussian copula construction for simulation,
* Matérn or exponential spatial correlation,
* log-link parameterization for Poisson fitting.

Everything else is bookkeeping, looping, and schema management.

---
