---
title: "H18-B: gcKrig Pipeline Example"
author: "LeBron James"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include = FALSE}
# This Rmd lives at the repo root (STAT778_Project/h18b_example.Rmd).
# Every source code file is under code/; outputs go to data/ and logs/.
# Nothing needs to be changed here because the working directory is
# already the repo root.

knitr::opts_chunk$set(
  message = TRUE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 6
)
```

## What this notebook does

Runs the H18-B generation pipeline interactively and inspects the
result. This is the exact same work the Slurm job does, but inside
RStudio where you can see each step.

---

## One-liner: run the full pipeline

```{r run_full_pipeline}
source("code/scripts/01_gen_h18b.R")
```

That's it. One file on disk: `data/generated/H18-B/default.rds`. Everything below is either inspecting that file or walking through the pipeline step by step for learning.

---

## Inspect the output

```{r inspect}
source("code/R/logging.R")
source("code/R/io.R")

s <- load_scenario("H18-B", "default")

str(s, max.level = 1)
```

The full input spec is preserved verbatim:

```{r config_roundtrip}
str(s$config, max.level = 2)
```

Summary of the simulated counts:

```{r summary}
values <- s$values[, 1]
summary(values)
cat("zeros:", sum(values == 0), "of", length(values), "\n")
```

---

## Visualize

```{r plot}
plot(
  s$locations[, 1], s$locations[, 2],
  pch = 19,
  cex = 1 + 2 * values / max(values),
  col = gray(1 - values / max(values)),
  xlab = "x", ylab = "y",
  main = "H18-B simulated counts (darker/bigger = higher)"
)
```

---

## Optional: step through the pipeline manually

If you want to see what happens at each stage:

```{r step_by_step, eval = FALSE}
library(gcKrig)

source("code/R/logging.R")
source("code/R/grids.R")
source("code/R/spec_utils.R")
source("code/R/sim_core.R")
source("code/R/io.R")
source("code/configs/h18b.R")

# Pick the paper target scenario
spec <- H18B_SCENARIOS$default

# 1. Build locations from the layout spec
locs <- do.call(make_locations, spec$layout)

# 2. Build gcKrig objects from the marginal + corr specs
mu_vec <- exp(spec$marginal$beta[1] +
              spec$marginal$beta[2] * locs[, 1] +
              spec$marginal$beta[3] * locs[, 2])

marginal_obj <- negbin.gc(mu = mu_vec, od = spec$marginal$od)
corr_obj     <- matern.gc(range  = spec$corr$range,
                          kappa  = spec$corr$kappa,
                          nugget = spec$corr$nugget)

# 3. Simulate (the wrapper handles seeding and logging)
scenario <- simulate_mc_scenario(
  spec      = spec,
  locations = locs,
  marginal  = marginal_obj,
  corr      = corr_obj
)

# 4. Save
save_scenario(scenario)
```

---

## Optional: try a variant without editing any file

```{r variant, eval = FALSE}
library(gcKrig)
source("code/R/logging.R")
source("code/R/grids.R")
source("code/R/spec_utils.R")
source("code/R/sim_core.R")
source("code/R/io.R")
source("code/configs/h18b.R")

variant <- h18b_spec("range_0.7", list(corr = list(range = 0.7)))

locs_v <- do.call(make_locations, variant$layout)
mu_v   <- exp(variant$marginal$beta[1] +
              variant$marginal$beta[2] * locs_v[, 1] +
              variant$marginal$beta[3] * locs_v[, 2])

scenario_v <- simulate_mc_scenario(
  spec      = variant,
  locations = locs_v,
  marginal  = negbin.gc(mu = mu_v, od = variant$marginal$od),
  corr      = matern.gc(range  = variant$corr$range,
                        kappa  = variant$corr$kappa,
                        nugget = variant$corr$nugget)
)

save_scenario(scenario_v)

list_scenarios("H18-B")  # should now show "default" and "range_0.7"
```

---

## What's next

- `sbatch code/slurm/01_gen_h18b.slurm` runs the same thing as a batch
  job on a Hopper compute node.
- To add a permanent variant, edit `code/configs/h18b.R` and append
  to `H18B_SCENARIOS` there.


---

## Deep dive: inspecting H18-B

### Load the scenario

```{r inspect_load}
source("code/R/logging.R")
source("code/R/io.R")

s <- load_scenario("H18-B", "default")
values <- s$values[, 1]              # n-vector of counts
locs   <- s$locations                # n x 2 matrix
```

### Scenario-level facts

```{r inspect_facts}
cat("Dataset:         ", s$dataset_id, "\n")
cat("Scenario:        ", s$scenario_id, "\n")
cat("Seed:            ", s$seed_info$seed, "\n")
cat("gcKrig version:  ", s$gcKrig_version, "\n")
cat("Created at:      ", format(s$created_at), "\n")
cat("n locations:     ", nrow(locs), "\n")
cat("n replicates:    ", ncol(s$values), "\n")
```

### Summary statistics of the simulated counts

```{r inspect_summary}
summary(values)

cat("\nvariance = ", var(values), "\n")
cat("variance / mean = ", var(values) / mean(values),
    "  (> 1 indicates overdispersion)\n")

cat("\nzeros:  ", sum(values == 0), " of ", length(values),
    " (", round(100 * mean(values == 0), 1), "%)\n", sep = "")
cat("max:    ", max(values), "\n")
```

Overdispersion should be well above 1 — the model has `od = 1`, meaning
the variance/mean ratio of the marginal distribution is $1 + \mu$, so
the overall empirical ratio should be substantially larger than the
Poisson value of 1.

### Distribution of counts

```{r inspect_hist, fig.width = 7, fig.height = 4}
hist(values,
     breaks = seq(-0.5, max(values) + 0.5, by = 1),
     col    = "gray80",
     border = "white",
     main   = "Distribution of simulated counts (H18-B)",
     xlab   = "count value")
```

A long right tail is expected — that's what "overdispersed" looks like
in practice.

### Spatial pattern: gradient + local smoothness

```{r inspect_spatial, fig.width = 7, fig.height = 6}
# Use count magnitude for both size and color intensity
size_scale  <- 0.5 + 2.5 * values / max(values)
color_scale <- gray(1 - values / max(values))

plot(
  locs[, "x"], locs[, "y"],
  pch  = 19,
  cex  = size_scale,
  col  = color_scale,
  xlab = "x coordinate",
  ylab = "y coordinate",
  main = "H18-B spatial pattern",
  xlim = c(-0.05, 1.05),
  ylim = c(-0.05, 1.05)
)

# Overlay the count values for clarity
text(locs[, "x"], locs[, "y"],
     labels = values,
     pos    = 3,            # above each point
     cex    = 0.6,
     col    = "red")
```

You should see two things:

1. A **trend**: higher counts in the upper-right (where $\mu(s)$ is
   largest), lower in the lower-left.
2. **Local smoothness**: nearby points tend to have similar counts,
   which is the spatial correlation at work. This is the whole point
   of the copula model — independent Poisson samples would not show
   this kind of clustering.

### Does the imposed trend actually appear?

```{r inspect_trend, fig.width = 9, fig.height = 4}
par(mfrow = c(1, 2))

plot(locs[, "x"], values,
     pch = 19, cex = 0.7, col = "gray40",
     xlab = "x coordinate", ylab = "count",
     main = "Count vs x")
abline(lm(values ~ locs[, "x"]), col = "red", lwd = 2)

plot(locs[, "y"], values,
     pch = 19, cex = 0.7, col = "gray40",
     xlab = "y coordinate", ylab = "count",
     main = "Count vs y")
abline(lm(values ~ locs[, "y"]), col = "red", lwd = 2)
```

Both slopes should be positive (the trend says so), and the $y$ slope should be steeper than the $x$ slope (since the coefficient on $y$ is twice that on $x$).

### The "truth" table

Since H18-B was simulated, we know the true parameters exactly. They live inside the scenario object for easy reference:

```{r inspect_truth}
s$true_params
```

When we later fit a model to this data with `mlegc()` in the analysis stage, the estimates should land close to these values. Estimates far from the truth would suggest either a bug or a genuinely hard inferential regime.

### Comparison with gcKrig's own plotter

`gcKrig` ships with a nice visualization for simulated data. We can
reconstruct a `simgc`-class object from our stored scenario and pass
it in:

```{r inspect_gckrig_plot, fig.width = 7, fig.height = 6}
library(gcKrig)

# Rebuild a simgc-compatible object
sim_like <- list(
  data = s$values,   # n x sim_n matrix
  locs = s$locations
)
class(sim_like) <- "simgc"

plot(sim_like, index = 1, plottype = "Dot", col = 2)
```

This should reproduce (approximately) Figure 1 middle panel from Han's
paper — the exact pattern differs because their example in that
figure uses a different seed and different parameters, but the
visual style is the same.